---
title: "argo bias"
subtitle: "analysis notebook"
author : daniel ross
date-modified: last-modified
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    theme: default
    toc: true
    toc-depth: 6
    toc-title: "contents"
    toc-location: right
    number-sections: true
lightbox: true
execute:
  echo: false
  warning: false
  message: false
  prompt: true
  fig-align: center
  fig-width: 11
  fig-height: 8.5
knitr:
  opts_chunk:
    prompt: false
    collapse: false
editor_options:
  chunk_output_type: console
---

# literature
## float velocity
- [K. Lebedev, H. Yoshinari, N. A. Maximenko, and P. W. Hacker. YoMaHa'07: Velocity data assessed from trajectories of Argo floats at parking level and at the sea surface, IPRC Technical Note No. 4(2), June 12, 2007, 16p.](https://apdrc.soest.hawaii.edu/projects/yomaha/)
  - derive Argo drift speed at park depth to quantify ocean circulation at 1,000 m depth
  - See also, [Katsumata, K., Yoshinari, H. Uncertainties in global mapping of Argo drift data at the parking level. J Oceanogr 66, 553–569 (2010). https://doi.org/10.1007/s10872-010-0046-4](https://link.springer.com/article/10.1007/s10872-010-0046-4)
  - Estimate parking level (1000m) zonal (eastward $u$) and meridional (northward $v$) velocity components using the coordinates ($[X,Y]$ = [lon, lat]) of floats before it dives and after it re-arrives at the sea surface
    \begin{gather}
      (U_\text{deep}^n, V_\text{deep}^n) = ([X,Y]^n_\text{first} - [X,Y]^{n-1}_\text{last}) / (T^n_\text{first} - T^{n-1}_\text{last})
    \end{gather}
  - Use a regression to estimate surface velocity and errors for deep/surface velocities
- Above referenced in more recent literature, e.g., [Oke, Peter R., Tatiana Rykova, Gabriela S. Pilo, and Jennifer L. Lovell. "Estimating Argo float trajectories under ice." Earth and Space Science 9, no. 7 (2022): e2022EA002312.](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022EA002312) 

# eda

- *analysis based on preprocessed `argo_object_01.RData` and is therefore limited to float observations in January of each year from 2007 through 2016*
- *augmented `preprocess_argo_data.R` to include velocity calculations*

**[[Note, just began incorporating by-year anaylses so these are incomplete]]**

```{r}
options(digits = 4, scipen = 999)

root <- getwd()
library(tidyverse)
library(data.table)
library(pbapply)
library(geosphere)
library(GpGp)

load(paste0(root, "/data/argo_velo_data_january.RData"))
```

## float preliminaries

- *drop observations for floats with only one observation*
- **velocity was calculated within year, therefore the first observation for a float in each year will have a missing velocity**
  - e.g., the velocity is not calculated between a given float's 2015-01-31 and 2016-01-01 profile
```{r}
argo[,.(
  floats = uniqueN(float_id),
  obs = .N,
  obs_velo = sum(!is.na(speed_ms)),
  dates = paste0(range(isodatetime), collapse = " - ")
), float_r]

A <- argo[(float_r)]
```

```{r}
oceans <- sort(unique(A$ocean))
oceans <- oceans[oceans!="south china and easter archipelagic seas"]
```

### obs per float
::: {.panel-tabset}
## overall
```{r}
float_n <- A[,.N,float_id]
summary(float_n$N)
hist(float_n$N, breaks = seq(0,44,1), main = "obs per float", xlab = "n")
```

## region
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  float_n <- A[ocean == o,.N,float_id]
  hist(float_n$N, breaks = seq(0,44,1), main = o, xlab = "n")
}
```
:::

### surfacing/obs frequency

- *most floats do share data every 10 days*

```{r}
A[, dt_day := dt_s / (60^2 * 24)]
summary(A$dt_day)
B <- A[!is.na(dt_day), .N, dt_day]

par(mfrow = c(1,1))
hist(A$dt_day, breaks = seq(0,31,.05), main = "obs freq", xlab = NULL)
#hist(A[dt_day < 50, dt_day], breaks = seq(0,50,.5), main = "0-50 days", xlab = NULL)
#hist(A[between(dt_day,300,400), dt_day], breaks = seq(300,400,1), main = "300-400 days", xlab = NULL)
#hist(A[between(dt_day,680,780), dt_day], breaks = seq(680,780,1), main = "680-780 days", xlab = NULL)

#b <- c(0,9,11,300,400,1000,2500)
#B[, cat := cut(dt_day, breaks = b, dig.lab = 5)]
#B[,.(obs = sum(N)),keyby = cat][,prop := obs / sum(obs)][, c_prop := cumsum(prop)][]
```

### ocean region 

- *for this analysis I used the global ocean regions*
- **[[update analysis to use finer ocean basin geographies. source?]]**
- *floats lat/lon reverse geocoded using shapefile available at <https://www.marineregions.org/sources.php#goas> and downloaded using the package `mregions2`*
- *dropped the 3 observations in the south china sea for regional analyses*


```{r}
float_loc <- A[,.(float_id, float_i, lon_degrees, lat_degrees, ocean, speed_ms)]

float_loc[,.(
  floats = uniqueN(float_id),
  obs = .N,
  obs_velo = sum(!is.na(speed_ms))
),ocean][order(-obs)]

sample_locs <- float_loc[, .SD[sample(.N, min(1000, .N))], by = ocean]

map_pts <- function(df, x = "lon_degrees", y = "lat_degrees", color = "ocean", size = .5, legend.rows = 1){
  ggplot() +
  borders("world", fill = "gray80", colour = "gray80") +
  geom_point(data = df,
    aes(x = .data[[x]],
        y = .data[[y]],
        color = .data[[color]]),
        size = size) +
  coord_sf() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = legend.rows))
}
map_sum_pts <- function(df, x = "lon_degrees", y = "lat_degrees", z = "u_ms", size = .5, legend.rows = 1){
  ggplot(df, aes(x = .data[[x]],y = .data[[y]],z = .data[[z]])) +
  borders("world", fill = "gray80", colour = "gray80") +
  stat_summary_2d(fun = mean, binwidth = c(1,1)) + 
  scale_fill_gradient2(midpoint = 0,low = "red",mid = "grey100",high = "blue") + 
  coord_sf() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_colorbar())
}

map_pts(sample_locs, color = "ocean", legend.rows = 3)
```

## velocity
- *velocity components u/v are calculated by finding the change in longitude/latitude and dividing by change in time.*
- *[`geosphere::distVincentyEllipsoid`](https://cran.r-project.org/web/packages/geosphere/refman/geosphere.html#distVincentyEllipsoid>), an ellipsoidal (WGS84 by default) great-circle-distance, is used to calculate the distance between float obs.*
- *also calculated the magnitude of the velocity (speed) and the bearing of the velocity vectors in radians*

### overall distributions
- *histograms of float velocity over entire ocean omitting extreme outliers corresponding to speeds >0.5m/s*
- *outliers are dropped for the remainder of this analysis*
```{r}
#| fig-height: 6
B <- A[!is.na(speed_ms)]

X <- A[!is.na(speed_ms) & speed_ms <= 0.5,]
X <- X[ocean != "south china and easter archipelagic seas"]
par(mfrow = c(1,3))
hist(X[,speed_ms], breaks = seq(0,.5,.01), main = "float speed", xlab = "speed (m/s)")
hist(X[,u_ms], breaks = seq(-.5,.5,.01), main = "float eastward velocity", xlab = "u (m/s)")
hist(X[,v_ms], breaks = seq(-.5,.5,.01), main = "float northward velocity", xlab = "v (m/s)")
```

```{r}
  ggplot(X, aes(x = u_ms, y = v_ms)) +
    stat_bin_2d(aes(fill = after_stat(density)), binwidth = 0.01) +
    scale_fill_viridis_c(option = "C") +
    labs(
      x = "u (m/s)",
      y = "v (m/s)",
      fill = "density"
    ) +
    theme_minimal() +
    coord_fixed()
```

```{r}
my.sum <- function(x) list(mean = mean(x), sd = sd(x))
X[, unlist(lapply(.SD, my.sum)), .SDcols = c("u_ms","v_ms","speed_ms")]
```

### outliers
- *for preliminary analyses simply dropped outliers and did not further analyze patterns in these obs*
```{r}
#b <- c(seq(0,.1,.01),seq(.2,.5,.1),1,10,20,50,100,500,1000)
b <- c(0,.5,1,10,20,50,100,500,B[,max(speed_ms)])
B[, speed_g := cut(speed_ms, breaks = b, dig.lab = 5)]
B[,.(obs = .N),keyby = speed_g][,prop := obs / sum(obs)][, c_prop := cumsum(prop)][]
```

### velocity maps

- *averaged speed/velocity within 1 deg by 1 deg cells*

::: {.panel-tabset}

## eastward velocity
```{r}
map_sum_pts(X, z = "u_ms") +
  labs(title = "u (m/s) eastward velocity")
```

## northward velocity
```{r}
map_sum_pts(X, z = "v_ms") +
  labs(title = "v (m/s) northward velocity")
```

## speed
```{r}
map_sum_pts(X, z = "speed_ms") + 
  labs(title = "speed (m/s)")
```

:::

### regional distributions

#### joint dist of velocity components
```{r}
ggplot(X, aes(x = u_ms, y = v_ms)) +
  stat_bin_2d(aes(fill = after_stat(density)), binwidth = 0.01) +
  facet_wrap(~ ocean, axes = "all") +
  scale_fill_viridis_c(option = "C") +
  labs(
    x = "u (m/s)",
    y = "v (m/s)",
    fill = "density"
  ) +
  theme_minimal() +
  coord_fixed()
```

```{r}
my.sum <- function(x) list(mean = mean(x), sd = sd(x))
X[, as.list(unlist(lapply(.SD, my.sum))), .SDcols = c("u_ms","v_ms","speed_ms"), keyby=ocean]
```

<!--
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    X[ocean == o, speed_ms],
    breaks = seq(0, 0.5, 0.01),
    main = o,
    xlab = "speed (m/s)"
  )
}
```

```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    X[ocean == o,u_ms],
    breaks = seq(-0.5, 0.5, 0.01),
    main = o,
    xlab = "u (m/s)"
  )
}
```

```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    X[ocean == o,v_ms],
    breaks = seq(-0.5, 0.5, 0.01),
    main = o,
    xlab = "v (m/s)"
  )
}
```
-->

## temperature

- *histograms of centered temperature by profile pressure (~depth in m)*
- *0m and 2000m correspond to the min/max observed pressure for each float*
- *for other depths, found the temperature associated with the pressure obs closest to the target depth*
  - *e.g., the most common pressure (8.4% of floats) for the '1000m' observations is in fact 988m, while 7.7% of floats have an obs at exactly 1000.0m*
- **for temperature only analyses, used data filtered to floats with multiple observations; however, included the first profile observation for those floats for each year where we have missing velocity **

<!-- X[,.N,pressure_1000][,prop := N / sum(N)][order(-N)]-->

::: {.panel-tabset}
### 0m
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_0_c],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 0m"
  )
}
```
### 100m
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_100_c],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 100m"
  )
}
```
### 500m
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_500_c],
   # freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 500m"
  )
}
```
### 1000m
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_1000_c],
   # freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 1000m"
  )
}
```
### 2000m
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_2000_c],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 2000m"
  )
}
```
### vhc
```{r}
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, vhc_obs],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "vhc"
  )
}
```
:::

### temperature maps

- *averaged temperature within 1 deg by 1 deg cells*

::: {.panel-tabset}
## 0m
```{r}
map_sum_pts(A, z = "temp_0") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_0)]) + 
  labs(title = "temp 0m")
```
## 100m
```{r}
map_sum_pts(A, z = "temp_100") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_100)]) + 
  labs(title = "temp 100m")
```
## 500m
```{r}
map_sum_pts(A, z = "temp_500") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_500)]) + 
  labs(title = "temp 500m")
```
## 1000m
```{r}
map_sum_pts(A, z = "temp_1000") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_1000)]) + 
  labs(title = "temp 1000m")
```
## 2000m
```{r}
map_sum_pts(A, z = "temp_2000") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_2000)]) + 
  labs(title = "temp 2000m")
```
## vhc 
```{r}
map_sum_pts(A, z = "vhc_obs") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(vhc_obs)]) + 
  labs(title = "vhc")
```
:::

## correlation between temperature and velocity

- *correlation between temperatures and velocities for floats with multiple observations*
- *dtemp_ variables are the change in temperature for consecutive float observations at the given depth*
  - negligible correlations between change in temperature and velocity
- **small to moderate negative correlation between eastward velocity and temperature/vhc across depths in southern hemisphere**
  - southern antarctic ocean which has positive correlation between eastward velocity and temperature/vhc
- **negligible negative correlation between eastward velocity and temperature/vhc across depths in northern hemisphere**
  - artic ocean has larger correlations but few obs
- **negligible or no correlation between northward velocity and temperature**

```{r}
#| echo: false
velo_temp_cor <- function(df, digits = 4){
  cols <- c("u_ms", "v_ms", "speed_ms",paste0("temp_",c(0,500,1000,2000)),"vhc_obs",paste0("dtemp_",c(0,500,1000,2000)),"dvhc_obs")
  V <- df[,..cols]
         return(round(cor(V)[-c(1:3),1:3],digits))
}
```

### excluding outlier velocities
#### overall
```{r}
#| results: asis
#| echo: false
cat("::: {.panel-tabset}\n\n")

cat("## overall", "\n\n")
tab <- knitr::kable(velo_temp_cor(X))
cat(tab, sep = "\n")
cat("N:", X[,.N], sep = "\n")
cat("\n")

for (o in oceans) {
  cat("##", o, "\n\n")
  tab <- knitr::kable(velo_temp_cor(X[ocean == o,]))
  cat(tab, sep = "\n")
  cat("N:", X[ocean == o,.N], sep = "\n")
  cat("\n")
  cat("\n\n")
}

cat(":::\n")
```

#### by year
::: {.panel-tabset}

```{r}
#| results: asis
#| echo: false

for (yr in 2007:2016){
  cat("##", yr, "\n\n")
  cat("::: {.panel-tabset}\n\n")

  cat("## overall", "\n\n")
  Y <- X[years == yr]
  tab <- knitr::kable(velo_temp_cor(Y))
  cat(tab, sep = "\n")
  cat("N:", Y[,.N], sep = "\n")
  cat("\n")

  for (o in oceans) {
    cat("##", o, "\n\n")
    tab <- knitr::kable(velo_temp_cor(Y[ocean == o,]))
    cat(tab, sep = "\n")
    cat("N:", Y[ocean == o,.N], sep = "\n")
    cat("\n")
    cat("\n\n")
  }
  cat(":::\n")
}
```

:::

<!--
#### by year
::: {.panel-tabset}
## 2016
```{r}
#| results: asis
#| echo: false
cat("::: {.panel-tabset}\n\n")

cat("## overall", "\n\n")
Y <- X[years == 2016]
tab <- knitr::kable(velo_temp_cor(Y))
cat(tab, sep = "\n")
cat("N:", Y[,.N], sep = "\n")
cat("\n")

for (o in oceans) {
  cat("##", o, "\n\n")
  tab <- knitr::kable(velo_temp_cor(Y[ocean == o,]))
  cat(tab, sep = "\n")
  cat("N:", Y[ocean == o,.N], sep = "\n")
  cat("\n")
  cat("\n\n")
}

cat(":::\n")
```
:::
-->

### including outlier velocities
- *overall, south atlantic, and southern correlations are sensitive to outliers*

```{r}
#| results: asis
#| echo: false

Y <- A[!is.na(speed_ms)]

cat("::: {.panel-tabset}\n\n")

cat("## overall", "\n\n")
tab <- knitr::kable(velo_temp_cor(Y))
cat(tab, sep = "\n")
cat("N:", Y[,.N], sep = "\n")
cat("\n")

for (o in oceans) {
  cat("##", o, "\n\n")
  tab <- knitr::kable(velo_temp_cor(Y[ocean == o,]))
  cat(tab, sep = "\n")
  cat("N:", Y[ocean == o,.N], sep = "\n")
  cat("\n")
  cat("\n\n")
}

cat(":::\n")
```

### heatmaps
- *centered temperatures averaged within 0.01m/s cells of (u, v)*
- *positive eastward velocity appears to be correlated with negative temperatures*

```{r}
plot_velocity_heatmap <- function(df,z,title = z,binwidth = 0.01, midpoint = 0, ocean = T) {

  p <- ggplot(df, aes(x = u_ms, y = v_ms, z = .data[[z]])) +
    stat_summary_2d(fun = mean, binwidth = binwidth) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
      midpoint = midpoint, name = "temp") +
    labs(x = "u (m/s)",y = "v (m/s)",title = title) +
    theme_minimal() +
    theme(legend.position = "bottom") +
    coord_equal()

  if(ocean){
    p <- p + facet_wrap(~ocean)
  }
  p
}
```

::: {.panel-tabset}
## 0m
```{r}
plot_velocity_heatmap(X, "temp_0_c", "centered ocean temp at 0m")
```
## 100m
```{r}
#plot_velocity_heatmap(X, "temp_100", "ocean temp at 100m", ocean=F)
```
```{r}
plot_velocity_heatmap(X, "temp_100_c", "centered ocean temp at 100m")
```
## 500m
```{r}
plot_velocity_heatmap(X, "temp_500_c", "centered ocean temp at 500m")
```
## 1000m
```{r}
plot_velocity_heatmap(X, "temp_1000_c", "centered ocean temp at 1000m")
```
## 2000m
```{r}
plot_velocity_heatmap(X, "temp_2000_c", "centered ocean temp at 2000m")
```
:::

## next steps
- include data from all months, not just january
- focus on finer ocean geographies
  - significant heterogeneity in temperature and velocity within large global ocean regions



<!--
```{r}
scatter <- function(df, x, y, main = NULL, line = TRUE){
  main <- ifelse(is.null(main), paste(x, "vs", y), main)
  plot(df[[x]], df[[y]],
     pch = 1, 
     xlab = x, 
     ylab = y,
     main = main
  )
  lines(lowess(df[[x]], df[[y]]), col = "red", lwd = 1)
}
```

```{r}
scatter(X, "u_ms", "dtemp_1000")
```

```{r}
scatter(X, "u_ms", "temp_1000")
```

```{r}
par(mfrow = c(2,4))
for(o in oceans){
  scatter(X[ocean==o], "u_ms", "dtemp_1000", main = o)
}
```

```{r}
par(mfrow = c(2,4))
for(o in oceans){
  scatter(X[ocean==o], "u_ms", "temp_1000", main = o)
}
```

```{r}
#NAO <- A[ocean %in% c("north atlantic ocean")]
#NAO[, has10 := any(between(dt_day,9,11)), float_id]
#NAO <- NAO[(has10),]
#map_pts(NAO, color = "speed_ms") + 
#  coord_sf(xlim = c(-80,5), ylim = c(0,65)) +
#  scale_color_viridis_c(option = "plasma", na.value = "grey80") +
#  guides(color = guide_colorbar())
#
#fid <- NAO[which.max(float_i), float_id]
#X <- NAO[float_id == fid]
#map_pts(X, color = "speed_ms") +
#  coord_sf(xlim = c(-80,5), ylim = c(0,65)) +
#  scale_color_viridis_c(option = "plasma", na.value = "grey80") +
#  guides(color = guide_colorbar())
```

```{r}
#X <- NAO[!is.na(speed_ms),.(speed_ms, u_ms, v_ms, z)]
#cor(X)
#cor.test(X$speed_ms, X$z)
#cor.test(X$u_ms, X$z)
#cor.test(X$v_ms, X$z)
```

```{r}
#S <- NAO[1:2, .(float_id, max_depth, obs_pressures, obs_temps, )]
```

```{r}
#fid <- A[which.max(float_i), float_id]
#X <- A[float_id == fid]
#map_pts(X, color = "speed_ms") +
#  scale_color_viridis_c(option = "plasma", na.value = "grey80") +
#  guides(color = guide_colorbar())
#X <- A[!is.na(speed_ms),.(speed_ms, u_ms, v_ms, z)]
#cor(X)
#ct <- cor.test(X$speed_ms, X$z)
#cor.test(X$u_ms, X$z)
#cor.test(X$v_ms, X$z)
```
-->

<!--
- calc. radius of drift (by region)
- do floats replace eachother in the same area? eg does a float move out of an initial radius but a different float moves into that initial radius?
  - within a fixed region (basins) how much area is covered by the floats over time? do they begin to cluster or do they circle and replace eachother such that the same area is covered?
- discuss distance metrics to calc. velocity
-->
