\section{Background}
\label{sec:background}

Argo floats park and drift at 1000m depth between profiling cycles.
Their locations are determined by currents at that depth, not by experimental design.
If 1000m currents are correlated with upper-ocean temperature or salinity, then Argo-based estimates of quantities like space-time averaged ocean heat content (OHC) could be biased.
This is widely suspected to be negligible, but the magnitude of this bias has not yet been bounded statistically.

This section sets up the estimand, the sampling mechanism, and the resulting sampling-induced bias.
Three logically distinct sources of randomness appear, each on its own probability space: a \emph{field} space carrying the ocean heat content (OHC) field, which---following model-based geostatistics and the hierarchical Bayesian OHC model of \citet{baugh2022} on which this work builds---we treat as a \emph{stochastic process}; a \emph{sampling} space carrying the Argo float point process; and a \emph{population} space on which a random space--time coordinate is drawn and against which the target average is defined.
Our analysis of the sampling bias is carried out \emph{conditional} on a realization of the field and on the sampling design; conditional on these, the field and the sampling weight are fixed functions and the only operative randomness is the target coordinate.
This conditioning is what allows a distribution-free bound to coexist with, and indeed validate, a stochastic-field model rather than contradict it.

\section{Methodology}
\label{sec:methodology}

All proofs are collected in Appendix~\ref{app:proofs}.

\subsection{Notation and Estimand}
\label{sec:def}

\begin{definition}[Space--time domain]
\label{def:domain}
Fix a spatial region of interest $A \subset S^{2}$, taken to be a Borel subset of the sphere with finite, positive surface area, and a study period $\T \subset \R$, a bounded interval.
The \emph{space--time domain} is the product set
$\X := A \times \T$, equipped with the product $\sigma$-algebra
$\F := \B(A)\otimes\B(\T)$, where $\B(A)$ and $\B(\T)$ are the Borel
$\sigma$-algebras on $A$ and $\T$. A generic element of $\X$ is written
$x = (s,t)$, with spatial location $s \in A$ and time $t \in \T$.
\end{definition}

\begin{definition}[Reference space--time measure]
\label{def:nu}
Let $\nu$ be the \emph{reference} or \emph{target} probability measure on $(\X,\F)$ such that $\nu(E) := |E|\, /\, |\X|$ for all  $E \in \F$.
\end{definition}

$\nu(E)$ is the probability that a profile drawn uniformly from the space-time domain lies in $E$.

\begin{definition}[Sampling probability measure]
\label{def:pi}
Let $\pi$ be the \emph{sampling} probability measure on $(\X,\F)$.
\end{definition}

$\pi(E)$ is the probability that a profile drawn at random from the \emph{sampling} or realized advection process lies in $E$. 

\begin{definition}[Target-population random point]
\label{def:X}
Let $X = (S,T)$ be the \emph{coordinate map} on $(\X,\F,\nu)$, i.e., the identity function $X : \X \to \X$, $X(x) = x$.
Its distribution is the reference measure by construction, $X \sim \nu$.
Thus, $\Prob(X \in E) = \nu(E)$ for every $E \in \F$. 
\end{definition}

\begin{definition}[Sampled-population random profile]
\label{def:X'}
Let $X' = (S,T)$ be a random variable on $(\X,\F,\pi)$, with distribution the sampling measure, $X' \sim \pi$.
Thus, $\Prob(X' \in E) = \pi(E)$ for every $E \in \F$. 
\end{definition}

\begin{definition}[Sampling weight]
\label{def:w}
Assume $\pi \ll \nu$. The \emph{sampling weight} is the Radon--Nikodym
derivative $w := \dd\pi/\dd\nu$: the $\nu$-almost-everywhere unique,
non-negative, $\F$-measurable function on $\X$ with
$\pi(E)=\int_E w\dd\nu$ for all $E\in\F$. We assume further that
$w \in L^2(\nu)$.
\end{definition}

Composing with the target draw $X\sim\nu$ (Definition~\ref{def:X}) gives
the non-negative random variable $w(X)$; note that $w:\X\to[0,\infty)$ is
a deterministic function on the domain, while $w(X)$ is its composition
with the random location, so that expectations such as $\E_\nu[w]$ below
are shorthand for moments of $w(X)$ under $X\sim\nu$. For $\nu$-almost
every $x\in\X$, Lebesgue differentiation gives
\[
  w(x) \;=\; \lim_{E \downarrow x} \frac{\pi(E)}{\nu(E)}
        \;=\; \lim_{E \downarrow x}
              \frac{\text{sampling mass of } E}{\text{reference mass of } E},
\]
so $w$ is the local ratio of sampling density to reference density. Where
$w \equiv 1$ the region is sampled exactly in proportion to its
space--time extent, i.e. as if uniformly; $w > 1$ marks over-sampling
(preferential), $w < 1$ under-sampling; and if $w = 0$ $\nu$-a.e. on a set
$E$ with $\nu(E)>0$, then $\pi(E)=0$ and $E$ is never sampled.

Because $\pi$ is a probability measure, $w(X)$ is mean-one,
\[
  \E_{\nu}\big[w(X)\big] \;=\; \int_{\X} w \dd\nu \;=\; \pi(\X) \;=\; 1,
\]
so $w-1$ is centered, and its dispersion under the reference measure,
\begin{equation}
  \sigma_w^2 \;:=\; \Var_{\nu}\big(w(X)\big)
  \;=\; \int_{\X} (w-1)^2 \dd\nu
  \;=\; \|w-1\|_{L^2(\nu)}^2
  \;=\; \int_{\X} w^2 \dd\nu - 1
  \;=\; \chi^2(\pi \,\|\, \nu),
  \label{eq:sigma-w}
\end{equation}
summarizes how inhomogeneous the sampling is: $\sigma_w = 0$ exactly when
$w = 1$ $\nu$-a.e., i.e. $\pi = \nu$ (non-preferential sampling), and
$\sigma_w$ grows as $\pi$ departs from $\nu$. Since $\E_\nu[w]=1$, the
scalar $\sigma_w$ is equivalently the coefficient of variation of $w(X)$,
a scale-free measure of sampling unevenness. Note that $w\in L^1(\nu)$ is
automatic, whereas the assumption $w\in L^2(\nu)$---equivalently
$\sigma_w<\infty$---is a genuine restriction, excluding sampling measures
so concentrated relative to $\nu$ that their $\chi^2$ divergence diverges.

Two features of \eqref{eq:sigma-w} shape what follows. It is defined
\emph{against the reference measure} $\nu$: it asks how far the sampling
law departs from the target-averaging weights. 

\begin{definition}[OHC random field]
\label{def:Y}
Let $Y: \Omega_Y \to L^2(\nu)$ be a random variable from ocean states to functional space.
\end{definition}

The bias we study is a property of the \emph{realized} ocean, so we condition on a realization of the field throughout.
Fix $\omega_Y\in\Omega_Y$ and write $y := Y(\omega_Y,\cdot)\in L^2(\nu)$ for the corresponding realization.
Because the study concerns a single physical ocean over the period $\T$, we identify $y$ with the true realized OHC field.
Conditional on the event $\{Y = y\}$; $y$ is a fixed element of $L^2(\nu)$.
Composing $y$ with the coordinate map yields the real-valued random variable $y(X): \X \to \R$ on the population space.

\begin{definition}[Estimand]
\label{def:mu}
The estimand is the \emph{realized mean OHC}, 
\[
  \mu = \E_\nu\big[y(X)\big] = \int_\X y \dd\nu.
\]
As a functional of the random field, $\mu(Y)$ is itself a random conditional on $\{Y=y\}$ it is a fixed scalar. 
Where
\[
  \sigma^2_Y(y) = \Var_{\nu}(y(X)) = \int_{\X}(y - \mu)^2\dd\nu = \|\,y - \mu\,\|^2_{L^2(\nu)}
\]
is the space--time variance of the realized field.
\end{definition}

\begin{definition}[Design mean]
\label{def:mu-w}
The \emph{design mean} is the realized field averaged under the
sampling measure,
\[
  \mu_w = \E_\pi\big[y(X')\big] = \int_\X y \dd\pi.
\]
Equivalenty,
\[
  \mu_w = \int_\X w\,y \dd\nu = \E_\nu\big[w(X)\,y(X)\big].
\]
So $w(X)$ reweights a target draw $X\sim\nu$ into a sampled draw $X'\sim\pi$.
Non-preferential sampling is $w\equiv 1$ ($\pi=\nu$).
\end{definition}

Let $\hat\mu$ denote any estimator of $\mu$ built from the observed profiles (for example a kriging/Gaussian-process interpolant).
Its bias decomposes into a sampling term and a reconstruction term.

\subsection{Bias Decomposition}
\label{sec:defbias}

\begin{definition}[Bias decomposition]
\label{def:bias}
The bias of $\hat\mu$ decomposes as
\[
  B(y) := \hat\mu - \mu
  = \underbrace{\big(\mu_w - \mu\big)}_{\textstyle B_{\mathrm{samp}}(y)}
  + \underbrace{\big(\hat\mu - \mu_w\big)}_{\textstyle B_{\mathrm{pred}}(y)},
\]
where $B_{\mathrm{samp}}(y)$ is the \emph{preferential-sampling bias}
and $B_{\mathrm{pred}}(y)$ is the \emph{prediction bias}. \end{definition}

The sampling term admits a covariance representation on the population
space, the starting point for the bound.

\begin{proposition}[Sampling-bias covariance identity]
\label{prop:cov-identity}
Conditional on $\{Y=y\}$ and the design, 
\[
  B_{\mathrm{samp}}(y) = \mu_w - \mu
  = \E_\nu\big[(w(X)-1)\,y(X)\big]
  = \Cov_\nu\big(w(X),\,y(X)\big).
\]
\end{proposition}
\deferproof{cov-identity}{%
\begin{proof}[Proof of Proposition~\ref{prop:cov-identity}]
By Definitions~\ref{def:mu} and~\ref{def:mu-w},
$\mu_w-\mu=\E_\nu[w(X)y(X)]-\E_\nu[y(X)]=\E_\nu[(w(X)-1)y(X)]$. Expanding the
covariance about $\mu$ and using~$\E_{\nu}[w(X)]=1$,
\begin{align*}
  \Cov_\nu(w(X),y(X)) &= \E_\nu\big[(w(X)-1)(y(X)-\mu)\big] \\
  &= \E_\nu\big[(w(X)-1)y(X)\big] - \mu\,\E_\nu[w(X)-1] \\
  &= \E_\nu\big[(w(X)-1)y(X)\big],
\end{align*}
since $\E_\nu[w(X)-1]=0$.
\end{proof}%
}

\begin{remark}[No sampling bias under uncorrelated design]
\label{rem:zero-bias}
If, for the realized field, the weight and the field are uncorrelated
under $\nu$---in particular whenever $w\equiv1$---then
$\Cov_\nu(w,y)=0$ and $B_{\mathrm{samp}}(y)=0$. Preferential-sampling
bias arises precisely from co-variation between where the floats
concentrate ($w$) and the value of the field ($y$).
\end{remark}

\begin{proposition}[Distribution-free $L^2$ bound]
\label{prop:l2-bound}
Under the conditions of Proposition~\ref{prop:cov-identity},
\[
  \big|B_{\mathrm{samp}}(y)\big|
  = \big|\Cov_\nu(w(X),y(X))\big|
  = |\rho_{w,y}|\,\sigma_w\,\sigma_y
  \;\le\; \sigma_w\,\sigma_y,
\]
where $\rho_{w,y}\in[-1,1]$ is the $\nu$-correlation between $w(X)$ and
$y(X)$. The bound requires only finite second moments and imposes no
smoothness assumption on the field.
\end{proposition}

\section{Estimation}
\label{sec:estimation}

\begin{definition}[Equal-measure partition and the projection operators]
\label{def:partition}
A \emph{level-$K$ partition} is a collection
$\mathcal{P}_K=\{C_1,\dots,C_K\}$ of disjoint measurable cells with
$\bigcup_j C_j=\X$ and equal reference measure $\nu(C_j)=1/K$ for all
$j$. Write $\PK:L^2(\nu)\to L^2(\nu)$ for the \emph{cell-averaging
operator}, i.e. the conditional expectation
$\PK f=\E_\nu[f\mid\mathcal{P}_K]$, which replaces a function by its cell
averages,
\[
  (\PK f)(x)=\bar f_j:=\frac{1}{\nu(C_j)}\int_{C_j} f\dd\nu
  \qquad\text{for }x\in C_j ,
\]
and write $\PKp:=\mathsf{I}-\PK$ for the complementary
\emph{within-cell} (or \emph{fluctuation}) operator,
\[
  (\PKp f)(x) := f(x)-(\PK f)(x) = f(x)-\bar f_j
  \qquad\text{for }x\in C_j .
\]
We abbreviate $f^{\perp}:=\PKp f$, so that every $f\in L^2(\nu)$ splits
as
\begin{equation}
  f=\PK f+\PKp f,\qquad\text{i.e.}\qquad \mathsf{I}=\PK+\PKp ,
  \label{eq:split}
\end{equation}
into its between-cell and within-cell parts.
\end{definition}

The two operators project onto complementary subspaces of $L^2(\nu)$.
Let
\[
  V_K=\Big\{\textstyle\sum_{j=1}^{K}a_j\mathbf{1}_{C_j}\;:\;a_j\in\R\Big\}
\]
be the $K$-dimensional subspace of functions constant on each cell (the
\emph{resolution-$K$} functions), whose indicator basis
$\{\mathbf{1}_{C_1},\dots,\mathbf{1}_{C_K}\}$ is orthogonal because the
cells are disjoint. Its orthogonal complement is
\[
  V_K^{\perp}=\Big\{g\in L^2(\nu)\;:\;\int_{C_j}g\dd\nu=0
    \ \text{for every }j\Big\},
\]
the functions with zero average on every cell, which carry only sub-cell
detail.

\begin{remark}
\label{rem:buildK}
Assuming $\nu$ is a measure over normalized area and time, equal-measure
cells are constructed by partitioning  
longitude into equal intervals, $\sin(\mathrm{lat})$ into equal
intervals (equal area on the sphere), and time into equal intervals,
so that $C_j$ is a longitude~$\times\sin(\mathrm{lat})\times$ time box
and $K=K_{\mathrm{lon}}K_{\mathrm{lat}}K_{\mathrm{t}}$. Each cell then
has equal meausre $\nu(C_j)=1/K$.
\end{remark}

\begin{lemma}[$\PK$ and $\PKp$ are complementary orthogonal projections]
\label{lem:projection}
For every $f\in L^2(\nu)$ we have $\PK f\in V_K$ and $\PKp f\in
V_K^{\perp}$; the operator $\PK$ is the orthogonal projection of
$L^2(\nu)$ onto $V_K$, and $\PKp$ the orthogonal projection onto
$V_K^{\perp}$. Consequently both are idempotent and self-adjoint,
\[
  \PK^2=\PK,\qquad (\PKp)^2=\PKp,\qquad \PK\PKp=\PKp\PK=0,
\]
their ranges are orthogonal, $\langle \PK f,\ \PKp g\rangle_\nu=0$ for all
$f,g\in L^2(\nu)$, and the Pythagorean identity
\begin{equation}
  \|f\|_{L^2(\nu)}^2=\|\PK f\|_{L^2(\nu)}^2+\|\PKp f\|_{L^2(\nu)}^2
  \label{eq:pythagoras}
\end{equation}
holds.
\end{lemma}
\deferproof{projection}{%
\begin{proof}[Proof of Lemma~\ref{lem:projection}]
$\PK f$ is constant on each cell, so $\PK f\in V_K$. For the residual,
use $\bar f_j\,\nu(C_j)=\int_{C_j}f\dd\nu$ from the definition of the
cell mean:
\[
  \int_{C_j}\PKp f\dd\nu
  =\int_{C_j}\big(f-\bar f_j\big)\dd\nu
  =\int_{C_j}f\dd\nu-\bar f_j\,\nu(C_j)=0
  \qquad\text{for every }j,
\]
so $\PKp f\in V_K^{\perp}$ by the characterization above. Orthogonality
of the ranges follows cell by cell: $\PK f$ takes the constant value
$\bar f_j$ on $C_j$, so
\[
  \big\langle \PK f,\ \PKp g\big\rangle_\nu
  =\sum_{j=1}^{K}\int_{C_j}\bar f_j\,(\PKp g)\dd\nu
  =\sum_{j=1}^{K}\bar f_j\underbrace{\int_{C_j}\PKp g\dd\nu}_{=\,0}
  =0 .
\]
In particular $f-\PK f$ is orthogonal to all of $V_K$, which is the
defining property of the orthogonal projection onto $V_K$; since
$\PKp=\mathsf{I}-\PK$, it is the orthogonal projection onto
$V_K^{\perp}$. Idempotence, self-adjointness and $\PK\PKp=0$ are the
standard consequences, and applying the orthogonality to
$f=\PK f+\PKp f$ gives \eqref{eq:pythagoras}.
\end{proof}%
}

Intuitively, subtracting the cell mean removes everything a
piecewise-constant function could capture: the average has already
extracted all of $f$ that is visible at cell resolution, so the residual
has nothing left to say there. That residual is $\PKp f$, the
\emph{within-cell fluctuation}: on each cell it equals $f$ minus that
cell's mean, and hence integrates to zero within every cell. It lives in
$V_K^{\perp}$, the functions carrying only sub-cell detail, so the split
$f=\PK f+\PKp f$ of~\eqref{eq:split} separates $f$ into its coarse,
between-cell part and its fine, within-cell part.

\subsection{Decomposing the weight and the field}
\label{sec:decomp-wy}

Lemma~\ref{lem:projection} applies to any $f\in L^2(\nu)$; the two
functions the bound requires are the centered weight $w-1$ and the field
anomaly $y-\mu$. Both are centered under $\nu$: $\E_\nu[w]=1$ by
\eqref{eq:sigma-w}, and $\E_\nu[y]=\mu$ by Definition~\ref{def:mu}.
Applying the operator identity~\eqref{eq:split} to each,
\begin{align}
  w-1 &= \underbrace{\PK(w-1)}_{\text{between-cell}}
        \;+\;\underbrace{\PKp(w-1)}_{\text{within-cell}}
        \;=\;w^{\star}+w^{\perp},
  \label{eq:w-split}\\[2pt]
  y-\mu &= \underbrace{\PK(y-\mu)}_{\text{between-cell}}
        \;+\;\underbrace{\PKp(y-\mu)}_{\text{within-cell}}
        \;=\;y^{\star}+y^{\perp},
  \label{eq:y-split}
\end{align}
where we have written
\[
  w^{\star}:=\PK(w-1)=\PK w-1,\qquad w^{\perp}:=\PKp(w-1)=\PKp w,
\]
\[
  y^{\star}:=\PK(y-\mu)=\PK y-\mu,\qquad y^{\perp}:=\PKp(y-\mu)=\PKp y,
\]
using that $\PK$ fixes constants ($\PK 1=1$, since $1\in V_K$) and hence
$\PKp$ annihilates them ($\PKp 1=0$). Concretely, the resolved parts
$w^{\star},y^{\star}$ are piecewise constant, equal to $\bar w_j-1$ and
$\bar y_j-\mu$ on $C_j$, while the unresolved parts $w^{\perp},y^{\perp}$
are $w-\bar w_j$ and $y-\bar y_j$ on $C_j$, each with zero average on
every cell.

\subsubsection*{Decomposition of the variances}

Because the two components in \eqref{eq:w-split} are orthogonal
(Lemma~\ref{lem:projection}), the Pythagorean
identity~\eqref{eq:pythagoras} splits the dispersion of the weight into
between- and within-cell pieces. Define
\begin{equation}
  \sigma_{w^{\star}} := \big\|w^{\star}\big\|_{L^2(\nu)},
  \qquad
  \sigma_{w^{\perp}} := \big\|w^{\perp}\big\|_{L^2(\nu)},
  \label{eq:sigma-w-parts}
\end{equation}
and analogously for the field,
\begin{equation}
  \sigma_{Y^{\star}} := \big\|y^{\star}\big\|_{L^2(\nu)},
  \qquad
  \sigma_{Y^{\perp}} := \big\|y^{\perp}\big\|_{L^2(\nu)} .
  \label{eq:sigma-Y-parts}
\end{equation}
Then
\begin{equation}
  \sigma_w^2=\sigma_{w^{\star}}^2+\sigma_{w^{\perp}}^2,
  \qquad
  \sigma_Y^2=\sigma_{Y^{\star}}^2+\sigma_{Y^{\perp}}^2 .
  \label{eq:var-split}
\end{equation}

These are exactly the law of total variance. Since
$\PK f=\E_\nu[f\mid\mathcal{P}_K]$, writing $X\sim\nu$ and letting
$J=J(X)$ denote the (random) index of the cell containing $X$,
\eqref{eq:var-split} reads
\[
  \underbrace{\Var_\nu\big(w(X)\big)}_{\sigma_w^2}
  =\underbrace{\Var_\nu\Big(\E_\nu\big[w(X)\mid J\big]\Big)}_{
      \sigma_{w^{\star}}^2\ =\ \sum_j \frac1K\,(\bar w_j-1)^2}
  \;+\;
  \underbrace{\E_\nu\Big[\Var_\nu\big(w(X)\mid J\big)\Big]}_{
      \sigma_{w^{\perp}}^2\ =\ \sum_j \frac1K\,\Var_\nu(w\mid C_j)} ,
\]
and identically for $y$: the resolved term is the variance of the cell
means, and the unresolved term is the mean of the within-cell variances.
In words, $\sigma_{w^{\star}}$ measures how unevenly sampling effort is
distributed \emph{across} cells---the dispersion of the cell weights
$\bar w_j=K\pi_j$ about one---while $\sigma_{w^{\perp}}$ measures the
sampling structure \emph{inside} cells, which the partition cannot see.
The same split separates field variation across cells from field
variation within them.

\subsubsection*{Why the split is useful}

The decompositions \eqref{eq:w-split}--\eqref{eq:y-split} are what allow
the bias itself to be separated by scale. Substituting them into the
covariance identity of Proposition~\ref{prop:cov-identity} and using the
orthogonality $\langle \PK f,\PKp g\rangle_\nu=0$ of
Lemma~\ref{lem:projection} to kill the two cross terms,
\begin{equation}
  B_{\mathrm{samp}}
  =\big\langle\, w-1,\; y-\mu \,\big\rangle_\nu
  =\underbrace{\big\langle\, w^{\star},\; y^{\star} \,\big\rangle_\nu}_{\text{between-cell}}
  \;+\;\underbrace{\big\langle\, w^{\perp},\; y^{\perp} \,\big\rangle_\nu}_{\text{within-cell}} ,
  \label{eq:bias-split}
\end{equation}
so the resolved and unresolved scales contribute additively and without
interference. Applying Cauchy--Schwarz to each term of
\eqref{eq:bias-split} separately pairs $\sigma_{w^{\star}}$ with
$\sigma_{Y^{\star}}$ and $\sigma_{w^{\perp}}$ with $\sigma_{Y^{\perp}}$,
giving
\[
  \big|B_{\mathrm{samp}}\big|
  \;\le\;\sigma_{w^{\star}}\,\sigma_{Y^{\star}}
  \;+\;\sigma_{w^{\perp}}\,\sigma_{Y^{\perp}} ,
\]
which is the content of Proposition~\ref{prop:scale-decomp}. This is what
makes a coarse partition---one the float counts can actually
support---admissible for bounding the bias, provided the within-cell
field variation $\sigma_{Y^{\perp}}$ is small.


\begin{proposition}[Scale-decomposition bound]
\label{prop:scale-decomp}
Conditional on the realized field $\{Y=y\}$ and the design, and under
the hypotheses of Proposition~\ref{prop:cov-identity}, for any
partition $\mathcal{P}_K$,
\[
  B_{\mathrm{samp}}
  =\underbrace{\big\langle\,w^{\star},\ y^{\star}\,\big\rangle_\nu}_{\text{between-cell}}
  +\underbrace{\big\langle\,w^{\perp},\ y^{\perp}\,\big\rangle_\nu}_{\text{within-cell}},
\]
and
\[
  |B_{\mathrm{samp}}|
  \;\le\; \sigma_{w^{\star}}\,\sigma_{Y^{\star}}
        \;+\;\sigma_{w^{\perp}}\,\sigma_{Y^{\perp}}
  \;\le\; \sigma_w\,\sigma_Y ,
\]
with $w^{\star},w^{\perp},y^{\star},y^{\perp}$ as in
\eqref{eq:w-split}--\eqref{eq:y-split} and the four dispersions as in
\eqref{eq:sigma-w-parts}--\eqref{eq:sigma-Y-parts}. In particular, the
scale-decomposed bound is never looser than the raw Cauchy--Schwarz
bound $\sigma_w\sigma_Y$.
\end{proposition}
\deferproof{scale-decomp}{%
\begin{proof}[Proof of Proposition~\ref{prop:scale-decomp}]
Substitute the decompositions $w-1=w^{\star}+w^{\perp}$ and
$y-\mu=y^{\star}+y^{\perp}$ of \eqref{eq:w-split}--\eqref{eq:y-split}
into the covariance identity
$B_{\mathrm{samp}}=\langle w-1,\,y-\mu\rangle_\nu$
(Proposition~\ref{prop:cov-identity}). Bilinearity of the inner product
gives four terms,
\begin{align*}
  B_{\mathrm{samp}}
  &=\big\langle\, w^{\star}+w^{\perp},\ y^{\star}+y^{\perp} \,\big\rangle_\nu\\[2pt]
  &=\underbrace{\big\langle\, w^{\star},\ y^{\star} \,\big\rangle_\nu}_{\text{between-cell}}
   \;+\;\underbrace{\big\langle\, w^{\star},\ y^{\perp} \,\big\rangle_\nu}_{\text{cross term}}
   \;+\;\underbrace{\big\langle\, w^{\perp},\ y^{\star} \,\big\rangle_\nu}_{\text{cross term}}
   \;+\;\underbrace{\big\langle\, w^{\perp},\ y^{\perp} \,\big\rangle_\nu}_{\text{within-cell}} .
\end{align*}
The two cross terms vanish by the orthogonality $V_K\perp V_K^{\perp}$ of
Lemma~\ref{lem:projection}: $w^{\star},y^{\star}\in V_K$ are constant on
each cell, while $w^{\perp},y^{\perp}\in V_K^{\perp}$ have zero average
on each cell. Explicitly, splitting the integral over the cells and
pulling out the constant value $\bar w_j-1$ of $w^{\star}$ on $C_j$,
\[
  \big\langle\, w^{\star},\ y^{\perp} \,\big\rangle_\nu
  =\int_{\X} w^{\star}\,y^{\perp}\dd\nu
  =\sum_{j=1}^{K}\int_{C_j}(\bar w_j-1)\,y^{\perp}\dd\nu
  =\sum_{j=1}^{K}(\bar w_j-1)\underbrace{\int_{C_j}y^{\perp}\dd\nu}_{=\,0}
  =0 ,
\]
and symmetrically $\langle w^{\perp},y^{\star}\rangle_\nu=0$, exchanging
the roles of $w$ and $y$. Only the between-cell and within-cell terms
survive, which is the stated two-term identity; Cauchy--Schwarz applied
to each of them separately yields
\[
  |B_{\mathrm{samp}}|
  \;\le\;\big|\big\langle w^{\star},y^{\star}\big\rangle_\nu\big|
        +\big|\big\langle w^{\perp},y^{\perp}\big\rangle_\nu\big|
  \;\le\;\sigma_{w^{\star}}\sigma_{Y^{\star}}
        +\sigma_{w^{\perp}}\sigma_{Y^{\perp}} .
\]

For the final inequality, apply the Cauchy--Schwarz inequality in $\R^2$
to the vectors
$\mathbf a=(\sigma_{w^{\star}},\sigma_{w^{\perp}})$ and
$\mathbf b=(\sigma_{Y^{\star}},\sigma_{Y^{\perp}})$:
\[
  \sigma_{w^{\star}}\sigma_{Y^{\star}}+\sigma_{w^{\perp}}\sigma_{Y^{\perp}}
  =\mathbf a\cdot\mathbf b
  \;\le\;\|\mathbf a\|\,\|\mathbf b\|
  =\sqrt{\sigma_{w^{\star}}^2+\sigma_{w^{\perp}}^2}\,
   \sqrt{\sigma_{Y^{\star}}^2+\sigma_{Y^{\perp}}^2}
  =\sigma_w\,\sigma_Y ,
\]
the last equality being the Pythagorean identities \eqref{eq:var-split}.
Equality holds precisely when $\mathbf a$ and $\mathbf b$ are parallel,
i.e. when $w$ and $y$ distribute their dispersion between the resolved
and unresolved scales in the same proportion,
$\sigma_{w^{\perp}}/\sigma_{w^{\star}}=\sigma_{Y^{\perp}}/\sigma_{Y^{\star}}$;
otherwise the scale-decomposed bound is strictly tighter.
\end{proof}%
}

\section{Implementation}
\label{sec:implementation}

The variance components that define the preferential sampling bound can be estimated without fitting models.
Namely, $\sigma_{w^{\star}}$ can be estimated from float counts (Section~\ref{sec:sigmawK}), and $\sigma_{Y^{\star}}$ and $\sigma_{Y^{\perp}}$ from a reanalysis product.
$\sigma_{w^{\perp}}$, the sampling variability \emph{within} the cell scale cannot be estimated from a sparse point pattern without reintroducing a within-cell intensity model.
However, proposition~\ref{prop:scale-decomp} confines it to a product with $\sigma_{Y^{\perp}}$, so the magnitude of $\sigma_{w^{\star}}$ only matters to the extent the OHC field varies within cells.
Therefore, our goal is to determine the cell scale such that within-cell OHC variability, $\sigma_{Y^{\perp}}$, is small such that the $\sigma_{w^{\star}}$ $\sigma_{Y^{\perp}}$ product is negligible.

\subsection{Choosing the resolution}
\label{sec:resolution}


The partition $\mathcal{P}_K$ is the only tuning choice in
\eqref{eq:sigmawK-hat}, and it is not free. The resolved dispersion
$\sigma_{w^{\star}}$ is genuinely resolution-dependent, and in a
particular direction.

\begin{lemma}[Coarsening is a monotone lower bound]
\label{lem:coarsening}
For any partition $\mathcal{P}_K$,
\[
  \sigma_{w^{\star}}\;\le\;\sigma_w,
  \qquad\text{with}\qquad
  \sigma_w^2-\sigma_{w^{\star}}^2
  =\sigma_{w^{\perp}}^2
  =\E_\nu\big[\Var_\nu(w\mid\mathcal{P}_K)\big]\;\ge\;0 .
\]
Moreover, under refinement the bound tightens monotonically: if
$\mathcal{P}_{K'}$ refines $\mathcal{P}_{K}$, in the sense that every
cell of $\mathcal{P}_{K}$ is a union of cells of $\mathcal{P}_{K'}$, then
$\sigma_{w^{\star}}(\mathcal{P}_{K})\le\sigma_{w^{\star}}(\mathcal{P}_{K'})\le\sigma_w$.
The same statements hold verbatim for the field, with
$(\sigma_{Y^{\star}},\sigma_{Y^{\perp}},\sigma_Y)$ in place of
$(\sigma_{w^{\star}},\sigma_{w^{\perp}},\sigma_w)$.
\end{lemma}

\deferproof{coarsening}{%
\begin{proof}
The identity is the Pythagorean decomposition \eqref{eq:var-split},
$\sigma_w^2=\sigma_{w^{\star}}^2+\sigma_{w^{\perp}}^2$, together with the
identification of the within-cell term as an expected conditional
variance (Section~\ref{sec:decomp-wy}); non-negativity of
$\sigma_{w^{\perp}}^2=\|w^{\perp}\|_{L^2(\nu)}^2$ gives
$\sigma_{w^{\star}}\le\sigma_w$. For monotonicity, refinement enlarges the
subspace of piecewise-constant functions, $V_{K}\subseteq V_{K'}$, and
the norm of an orthogonal projection cannot decrease when the range is
enlarged: $\|\mathsf{P}_{V_{K}}f\|\le\|\mathsf{P}_{V_{K'}}f\|$ for every
$f\in L^2(\nu)$. Applying this to $f=w-1$ gives the claim.
\end{proof}%
}

Two consequences shape everything below. First, the coarse estimator is
\emph{conservative in the wrong direction} for a naive bound: since
$\sigma_{w^{\star}}\le\sigma_w$, the product $\sigma_{w^{\star}}\sigma_Y$
is \emph{not} a valid upper bound on $|B_{\mathrm{samp}}|$---it may
understate it. This is precisely why the bound must be taken in its
scale-decomposed form,
$\sigma_{w^{\star}}\sigma_{Y^{\star}}+\sigma_{w^{\perp}}\sigma_{Y^{\perp}}$
(Proposition~\ref{prop:scale-decomp}), which pairs the coarse weight
dispersion only with the coarse field dispersion and accounts separately
for what lies below the cell scale.

Second, $K$ cannot be chosen by refining until the estimate stabilizes.
Lemma~\ref{lem:coarsening} says the refinement curve
$K\mapsto\sigma_{w^{\star}}$ climbs monotonically toward $\sigma_w$; and
because Argo trajectories cluster at short range, the empirical curve
does not plateau within any resolution the data support---it simply
continues to rise, at fine scales driven by trajectory autocorrelation
and count noise rather than by preferential sampling
(Section~\ref{sec:resolution}, ``Why the estimator degrades when
$N/K<1$''). There is no resolution at which the data reveal ``the true
$\sigma_w$''. The cell scale must therefore be fixed \emph{externally},
and the remainder of this section does so by bracketing it between two
limits: a \emph{relevance} scale from above, below which resolving $w$ is
unnecessary, and a \emph{sampling} scale from below, beyond which
resolving $w$ is infeasible. Three scales are in play and must not be
conflated.

The lemma applies to the field as well, but with a decisive difference in
consequence. For the weight, $\sigma_{w^{\star}}\le\sigma_w$ is a genuine
obstruction: the missing piece $\sigma_{w^{\perp}}$ cannot be recovered
from a sparse point pattern, and the refinement curve does not plateau,
so $\sigma_w$ is not identified from the data. For the field, by
contrast, the reanalysis is dense: $\sigma_{Y^{\star}}$,
$\sigma_{Y^{\perp}}$ and $\sigma_Y$ are all directly computable on the
native grid, and the identity
$\sigma_Y^2=\sigma_{Y^{\star}}^2+\sigma_{Y^{\perp}}^2$ can be verified
numerically. Of the four factors in the scale-decomposed bound of
Proposition~\ref{prop:scale-decomp}, exactly one---$\sigma_{w^{\perp}}$---is
unavailable, and it is paired with one---$\sigma_{Y^{\perp}}$---that is
both measurable and, by Lemma~\ref{lem:coarsening}, monotonically
decreasing under refinement. This pairing is what makes the bound usable:
refining the partition drives $\sigma_{Y^{\perp}}\downarrow 0$ and with it
the unestimated product, at the cost of driving $N/K$ toward the
coincidence-starved regime in which $\sigma_{w^{\star}}$ itself becomes
unestimable. The two monotonicities run in opposite directions along the
same refinement axis, and the operating band is where both remain
tolerable.

%The partition $\mathcal{P}_K$ is the only tuning choice in \eqref{eq:sigmawK-hat}, and it is not free: $\sigma_{w^{\star}}$ is genuinely resolution-dependent, increasing monotonically to $\sigma_w$ as the partition refines (Lemma~\ref{lem:coarsening}).
%It cannot therefore be selected by refining until the estimate stabilizes---there is no plateau to find, and for a finite point pattern the empirical curve simply climbs.
%The cell scale must be fixed \emph{externally}, and this section does so by bracketing it between two physically and statistically meaningful limits: a \emph{relevance} scale from above, below which resolving $w$ is unnecessary, and a \emph{sampling} scale from below, beyond which resolving $w$ is infeasible.
%Three scales are in play and must not be conflated.

\subsubsection*{The rendering scale}

The reanalysis is delivered at $1/12^\circ$ (${\approx}7$\,km here). This
is the resolution at which the \emph{field product} carries information,
and it is the right grid on which to compute the field factors
$\sigYhat$ and $\sigYperphat$ internally via
\eqref{eq:sigmaYK-hat}--\eqref{eq:sigmaYperp-hat}---in particular,
$\sigma_{Y^{\perp}}$ must be evaluated from the native grid rather than
from cell-averaged values, or the sub-cell variance it is meant to
capture is discretized away. It says nothing, however, about the
resolution the float pattern can support, and it is not a candidate cell
scale for $\sigma_{w^{\star}}$.

\subsubsection*{The relevance scale: how finely $w$ needs to be resolved}

By Proposition~\ref{prop:scale-decomp}, the one factor we decline to
estimate, $\sigma_{w^{\perp}}$, enters the bound only through the product
$\sigma_{w^{\perp}}\sigma_{Y^{\perp}}$. Sub-cell sampling structure is
therefore bias-irrelevant to the extent that the field is constant within
cells: choosing cells no larger than the OHC decorrelation length makes
$\sigma_{Y^{\perp}}$ small and suppresses the unestimated term regardless
of how rough the sub-cell sampling turns out to be. The relevance scale
is thus the OHC decorrelation length, and we fix it empirically from the
reanalysis rather than by inspection.

\paragraph{Procedure.} For each estimand (raw OHC and deseasonalized
OHC; Section~\ref{sec:sigmaY}) we proceed as follows.
\begin{enumerate}
\item \emph{Field construction.} Compute OHC on the native $1/12^\circ$
  grid over the analysis window. For the raw estimand, remove a
  per-snapshot planar trend in $(\mathrm{lon},\mathrm{lat})$; the box
  spans the Gulf Stream front, whose standing gradient otherwise renders
  the variogram non-stationary and unbounded. For the deseasonalized
  estimand, subtract the reanalysis climatology
  (Remark~\ref{rem:glorys}) and remove the per-snapshot mean.
\item \emph{Projection.} Convert $(\mathrm{lon},\mathrm{lat})$ to a local
  planar frame in kilometres about the box centre, so that lags and
  ranges are directly comparable to the sampling scale of
  \eqref{eq:cell-km}.
\item \emph{Directional empirical variogram.} Within each snapshot, form
  all pairs of grid points (subsampling for tractability), and accumulate
  the semivariance $\tfrac12(y_i-y_j)^2$ into bins of separation
  distance. Classify each pair by orientation and accumulate separately
  for zonal ($|\theta|<30^\circ$), meridional ($|\theta|>60^\circ$) and
  omnidirectional bins. Pool across snapshots. \emph{The variogram must
  be directional}: the frontal structure of this region is strongly
  anisotropic, with long correlation along the current and short
  correlation across it, and an isotropic average would conceal the
  short scale, which is the binding one.
\item \emph{Range extraction.} From each directional variogram read the
  decorrelation length two ways---the empirical $1/e$ crossing, i.e.\ the
  lag at which $\gamma(h)=(1-e^{-1})\,\mathrm{sill}$, and the range
  parameter of a fitted exponential (or Mat\'ern) model---and report
  both. The $1/e$ crossing is the robust primary, as the parametric fit
  can fail on noisy or hole-effect variograms.
\item \emph{Temporal scale.} Separately, compute the lagged
  autocorrelation of the de-seasonalized regional-mean OHC series and
  read its $e$-folding lag, which together with the seasonal amplitude
  determines the number of time bins $K_t$
  (Section~\ref{sec:leakage}).
\end{enumerate}

\paragraph{The binding number.} Write $L_{\mathrm{merid}}$ and
$L_{\mathrm{zonal}}$ for the two directional ranges. Because
$\sigma_{Y^{\perp}}$ is controlled by field variation in \emph{every}
direction within a cell, it is the \emph{shorter} of the two---here the
across-front, meridional scale---that governs the residual term. The
relevance scale is therefore $L_{\mathrm{rel}}:=L_{\mathrm{merid}}$, and
the requirement is that cells be no coarser than $L_{\mathrm{rel}}$. As
the ranges depend on the de-trending choices of step~1, we report them
under an alternative de-trending as a sensitivity check and, where these
disagree, adopt the shorter (conservative) value.

\subsubsection*{The sampling scale: how finely $w$ can be resolved}

For a $K_{\mathrm{lon}}\times K_{\mathrm{lat}}$ partition of the box---%
here $\Delta\lambda=6^\circ$ of longitude by $\Delta\phi=4^\circ$ of
latitude about $\phi_0=38^\circ$---the zonal and meridional cell sides
are
\begin{equation}
  \Delta x=\frac{\Delta\lambda}{K_{\mathrm{lon}}}\,(111\,\mathrm{km})\cos\phi_0
    =\frac{525\,\mathrm{km}}{K_{\mathrm{lon}}},
  \qquad
  \Delta y=\frac{\Delta\phi}{K_{\mathrm{lat}}}\,(111\,\mathrm{km})
    =\frac{444\,\mathrm{km}}{K_{\mathrm{lat}}},
  \label{eq:cell-km}
\end{equation}
taking $111$\,km per degree and $\cos38^\circ\approx0.79$. Two point
counts bracket the feasible band, each converted to a spacing through
$s=\sqrt{|A|/n}$---the side of one of $n$ equal cells tiling the domain
area $|A|\approx2.3\times10^{5}\,\mathrm{km}^2$
(Figure~\ref{fig:sampling-scale}). All $N=364$ profiles give
$s\approx25$\,km, i.e.\ ${\approx}21\times18$ cells: the
one-profile-per-cell scale, $K\approx N$. The $F=40$ distinct floats---%
the independent replication units---give $s\approx76$\,km, i.e.\
$K_{\mathrm{lon}}\times K_{\mathrm{lat}}\approx7\times6$.

\begin{figure}[t]
\centering
\begin{tikzpicture}[x=0.62cm,y=0.62cm]
  \def\cols{6}
  \def\rows{5}
  \fill[blue!12] (0,0) rectangle (1,1);
  \draw[line width=0.9pt] (0,0) rectangle (\cols,\rows);
  \foreach \i in {0,...,5}{\foreach \j in {0,...,4}{%
    \fill (\i+0.5,\j+0.5) circle (1.1pt);}}
  \draw[line width=0.9pt] (0,0) rectangle (1,1);
  \draw[<->] (0,-0.5) -- (1,-0.5) node[midway,below]{$s$};
  \draw[<->] (-0.5,0) -- (-0.5,1) node[midway,left]{$s$};
  \node[above] at (\cols/2,\rows+0.15) {$A$};
\end{tikzpicture}
\caption{The sampling scale $s=\sqrt{|A|/n}$. Tiling the analysis domain
(area $|A|$) into $n$ equal cells makes each a square of side $s$ with
$n\,s^{2}=|A|$; equivalently $s$ is the spacing of $n$ points placed one
per cell (dots). For our box,
$|A|\approx2.3\times10^{5}\,\mathrm{km}^2$, the $F=40$ floats give
$s\approx76$\,km ($K_{\mathrm{lon}}\times K_{\mathrm{lat}}\approx7\times6$)
and all $N=364$ profiles give $s\approx25$\,km (${\approx}21\times18$)---%
the coarse and fine ends of the resolution band.}
\label{fig:sampling-scale}
\end{figure}

\paragraph{Why the estimator degrades when $N/K<1$.} The mean occupancy
$N/K$ is the natural measure of how much the counts can support, and
$N/K<1$ marks a qualitative change rather than merely a noisier regime.
Estimator~\eqref{eq:sigmawK-hat} reads the data only through
coincidences: by \eqref{eq:ustat}, $\widehat{\sigma_{w^{\star}}^2}=K U_N-1$
where $U_N$ is the fraction of distinct profile pairs sharing a cell, and
under near-uniform sampling the expected number of coincident ordered
pairs is
\[
  \E\Big[\sum_j n_j(n_j-1)\Big]=N(N-1)\sum_j\pi_j^2\approx\frac{N^2}{K}
  =\frac{N}{K}\cdot N .
\]
Once $N/K<1$ this total falls below $N$ and is often in the single
digits: cells holding a lone profile contribute $n_j(n_j-1)=0$, so the
statistic is determined by the few cells that happen to catch two. Four
consequences follow.

\emph{Quantization.} One additional coincident pair moves
$\widehat{\sigma_{w^{\star}}^2}$ by $2K/[N(N-1)]\approx 2K/N^2$, which
exceeds $2/N$ once $N/K<1$. The estimator's support becomes a coarse
lattice and it cannot resolve small departures from uniformity even in
principle.

\emph{Variance inflation.} Near uniformity $\Var(U_N)$ is of order
$\theta/N\approx 1/(NK)$, so
$\Var\big(\widehat{\sigma_{w^{\star}}^2}\big)\sim K/N$: the standard error
grows like $\sqrt{K/N}$ while the estimand remains $O(1)$. That this is
the same ratio $K/N$ governing the plug-in bias of
Lemma~\ref{lem:unbiased} is no accident---both pathologies have the
single source of too few pairs per cell.

\emph{Grid-phase dependence.} At fine scales, whether two nearby profiles
share a cell depends on where the boundaries happen to fall. Two
partitions with identical $K$ can yield materially different
$\widehat{\sigma_{w^{\star}}}$, so the statistic reads boundary placement
rather than sampling structure---the signature of having passed the
resolvable scale.

\emph{Degeneracy and untrustworthy intervals.} When no two distinct
profiles share a cell, $\sum_j n_j(n_j-1)=0$ and
$\widehat{\sigma_{w^{\star}}^2}=-1$ exactly, its minimum; this occurs with
certainty once $K\gtrsim N^2$. The delete-one-float jackknife
\eqref{eq:jackvar} is consistent for smooth functionals, but in this
regime the statistic is a jumpy, discrete function of the data and the
justification lapses---indeed the jackknife standard error collapses
toward zero where the estimate is degenerate, so intervals can
spuriously \emph{narrow} exactly where they are least reliable.

Accordingly we require mean occupancy comfortably above one. This is the
same guidance as the classical requirement that Pearson expected cell
counts exceed ${\approx}5$, and for the same reason:
$\widehat{\sigma_{w^{\star}}^2}^{\,\mathrm{plug}}=X^2/N$ exactly, so the
sparse-cell caveat transfers verbatim. Because the independent units are
the $F$ floats rather than the $N$ profiles, the effective occupancy is
nearer $F/K$ than $N/K$ and the constraint binds sooner than the raw
profile count suggests; the float-based spacing $s\approx76$\,km is
therefore the honest coarse anchor.

\subsubsection*{The operating band, and the verdict}

The two scales bracket the choice from opposite directions: cells must be
\emph{fine enough} that $\sigma_{Y^{\perp}}$ is small
($\Delta x,\Delta y\lesssim L_{\mathrm{rel}}$), and \emph{coarse enough}
that coincidences are countable ($N/K$, effectively $F/K$, not below
one). The band is non-empty precisely when
\begin{equation}
  L_{\mathrm{rel}} \;\gtrsim\; s_F=\sqrt{|A|/F}\approx 76\,\mathrm{km},
  \label{eq:verdict}
\end{equation}
i.e.\ when the field's shortest decorrelation length is no finer than the
scale the floats can resolve. Condition \eqref{eq:verdict} is not assumed:
it is the verdict that the variogram procedure above returns, and it is
reported as such. Its two branches lead to materially different claims,
and which obtains is an empirical matter to be settled per region and
period.

If \eqref{eq:verdict} holds, cells at the sampling scale sit within a
decorrelation length, $\sigma_{Y^{\perp}}$ is correspondingly small, and
the residual $\sigma_{w^{\perp}}\sigma_{Y^{\perp}}$ is negligible against
the leading term $\sigma_{w^{\star}}\sigma_{Y^{\star}}$. The bound is then
certified at the operating resolution: the sampling structure we cannot
resolve lies at scales where the field does not vary, and so cannot bias
a regional mean.

If instead $L_{\mathrm{rel}}<s_F$, there is a band of field structure,
between $L_{\mathrm{rel}}$ and $s_F$, that is bias-relevant but that the
float pattern cannot resolve. The residual term is then not automatically
negligible and must be confronted rather than assumed away: we report the
bound as conditional on the unresolved $\sigma_{w^{\perp}}$, rather than
claiming a certified envelope.

In either branch we report the diagnostic
\begin{equation}
  \frac{\sigYperphat}{\hat\sigma_Y}
  \;=\;\text{fraction of field variance below the cell scale},
  \label{eq:residual-diagnostic}
\end{equation}
computed from the reanalysis on the operating partition, as direct
evidence of the magnitude of the dropped term. Because $\sigma_{Y^{\perp}}$
is estimable from the dense reanalysis even when $\sigma_{w^{\perp}}$ is
not, \eqref{eq:residual-diagnostic} quantifies how much of the residual
product can be bounded regardless of the verdict---a small value confines
the unresolved sampling roughness to scales at which it cannot matter,
whatever its size. This test, applied to any region and period, is what
determines whether a bound can be certified there, and it is the honest
failure mode of the method: where the data cannot support a certified
bound, the procedure says so rather than returning a reassuring number.

\subsubsection*{Reporting}

Because $\sigma_{w^{\star}}$ is resolution-dependent by construction, we
report not a single value but the \emph{refinement curve}
$K\mapsto\sigwhat$ across the $25$--$76$\,km band, each rung carrying its
float-clustered interval (Section~\ref{sec:uncertainty}), with
$L_{\mathrm{rel}}$ and $s_F$ marked. The field factors
$\sigYhat,\sigYperphat$ are evaluated on the \emph{same} partition at
each rung---computed on the $1/12^\circ$ grid and aggregated to the
current cells---so that the two fields share the partition at every
resolution, which is the only consistency
Proposition~\ref{prop:scale-decomp} requires.

Within this band we evaluate the all-pairs
estimator~\eqref{eq:sigmawK-hat} on all $N=364$ profiles rather than
retreating to the coarse, float-limited end. Resting on more pairs and
better-populated cells, it has lower sampling variance than its
between-float counterpart~\eqref{eq:sigmawK-bf}; and where consecutive
surfacings of a float re-enter the same cell, the additional
within-float coincidences inflate the estimate, which by
Lemma~\ref{lem:clustering-inflates} is the conservative direction for a
bound (under the checkable condition $c\ge\theta$ that a float coincides
with itself at least as often as two independent draws would). Its
\emph{independent} information nonetheless amounts to only $F=40$
floats, so uncertainty is quantified by float-level resampling rather
than by treating profiles as independent, and we report
\eqref{eq:sigmawK-bf} alongside as the estimate free of any within-float
independence assumption. The gap between the two is itself the diagnostic
of how much apparent dispersion is trajectory autocorrelation rather than
preferential sampling.


\subsection{Estimating $\sigma_{w^{\star}}$ from float counts}
\label{sec:sigmawK}

\begin{definition}[Cell sampling weight]
\label{def:sweight}
Let $\bar w_j$ be the \emph{sampling weight} for cell $j$, which is the ratio of the sampling mass of the cell to its reference mass,
\begin{equation}
  \bar w_j \;=\; \frac{1}{\nu(C_j)}\int_{C_j}w\dd\nu \;=\; \frac{\pi(C_j)}{\nu(C_j)} \;=\; \frac{\pi_j}{1/K}
  \;=\; K\pi_j ,
  \label{eq:cell-weight}
\end{equation}
where $\pi_j:=\pi(C_j)$ for the probability that a randomly sampled profile lies in cell $C_j$.
\end{definition}

This is coarse cell counterpart of the pointwise weight $w$ of
Definition~\ref{def:w}: $\bar w_j=1$ means cell $j$ is sampled exactly as
under uniform coverage, $\bar w_j>1$ over-sampled (preferential),
$\bar w_j<1$ under-sampled, and $\bar w_j=0$ unsampled. Because
$\sum_j\pi_j=1$ and $\nu(C_j)=1/K$, the weights are mean-one under the
reference measure, $\sum_j\nu(C_j)\bar w_j=\sum_j\pi_j=1$.

Let $n_j:=N(C_j)$ be the number of profiles observed in cell $j$ and
$N=\sum_j n_j$ the total. The natural estimator of $\bar w_j$ is
\begin{equation}
  \widehat{\bar w}_j=K\,\frac{n_j}{N}=\frac{n_j}{\,N/K\,}
  =\frac{\text{observed count in }C_j}{\text{count expected under uniform}} ,
  \label{eq:wbar-hat}
\end{equation}
so that a map of $\widehat{\bar w}_j$ over the domain displays directly
where floats over- and under-sample relative to uniform coverage.

\begin{lemma}[Cell sampling weight variance]
\label{lem:sigmawcell}
The variance of the cell sampling weights is
\[
  \sigma^2_{w^{\star}} = K\sum_{j=1}^{K}\pi_j^2-1 .
\]
\end{lemma}
\deferproof{sigmawcell}{%
\begin{proof}[Proof of Lemma~\ref{lem:sigmawcell}]
$\sigma^2_{w^{\star}}$ is the variance of the cells weights about one.
This is because $w^{\star}=\PK w-1$ is a constant function on each cell with cell value on $C_j$ is $\bar w_j-1$.

Accordingly,
\begin{align*}
  \sigma_{w^{\star}}^2
  &=\int_{\X}(w^{\star})^2\dd\nu
   =\sum_{j=1}^{K}\int_{C_j}\big(\bar w_j-1\big)^2\dd\nu
   =\sum_{j=1}^{K}\big(\bar w_j-1\big)^2\,\nu(C_j)
   =\frac{1}{K}\sum_{j=1}^{K}\big(K\pi_j-1\big)^2 \\
  &=\frac{1}{K}\sum_{j=1}^{K}\Big(K^2\pi_j^2-2K\pi_j+1\Big)
   =K\sum_{j=1}^{K}\pi_j^2\;-\;2\underbrace{\sum_{j=1}^{K}\pi_j}_{=\,1}
     \;+\;\frac{1}{K}\underbrace{\sum_{j=1}^{K}1}_{=\,K} \\
  &=K\sum_{j=1}^{K}\pi_j^2-1 .
  \label{eq:sigmawK-def}
\end{align*}
\end{proof}%
}

Note, uniform sampling within the partition ($\pi_j=1/K$ for all $j$, i.e.
$\bar w_j\equiv1$) gives $K\cdot K^{-1}-1=0$.
Whereas when all sampling mass is in a single cell gives $K-1$.
Hence, $0\le\sigma_{w^{\star}}\le\sqrt{K-1}$.
Thus, between-cell sampling variability can increase arbitrarily as the number of cells $K$ grows, degenerating as almost all cells have zero observed counts.

\begin{assumption}[Profiles sample the sampling measure]
\label{ass:sample-pi}
Conditional on the total $N$, the profile locations are independent draws
from the sampling measure $\pi$ of Definition~\ref{def:pi}; equivalently,
the cell counts follow
$(n_1,\dots,n_K)\sim\mathrm{Multinomial}\big(N;\pi_1,\dots,\pi_K\big)$
with $\pi_j=\pi(C_j)$.
\end{assumption}

The plug-in $\hat\pi_j=n_j/N$ gives
\[
  \sigwhatsqplug
  =K\sum_j\hat\pi_j^2-1=\frac{X^2}{N},
  \qquad X^2=\sum_{j=1}^{K}\frac{(n_j-N/K)^2}{N/K},
\]
where $X^2$ is \emph{Pearson's goodness-of-fit statistic for
uniformity}.
This plug-in estimator is upward biased and removing the diagonal (self-pairs) removes the bias (Lemma~\ref{lem:unbiased}):
\begin{equation}
  \sigwhatsq
  \;=\;K\sum_{j=1}^{K}\frac{n_j(n_j-1)}{N(N-1)}\;-\;1
  \label{eq:sigmawK-hat}
\end{equation}

\begin{remark}[Negative estimates]
\label{rem:negative}
Being unbiased for a non-negative quantity,
\eqref{eq:sigmawK-hat} can take negative values in finite samples.
A negative value arises exactly when the observed coincidence rate falls below the uniform rate, $\sum_j n_j(n_j-1)/[N(N-1)]<1/K$, and so indicates sampling at least as
even as the reference measure.
We retain the signed value for interval construction and report
$\sigwhat=\sqrt{\max\{0,\sigwhatsq\}}$
as the point estimate; near-zero results are stated on the $\sigma^2$
scale with their signed interval, as ``consistent with zero'' rather than
truncated.
\end{remark}

\begin{lemma}[Bias of the plug-in; unbiasedness of \eqref{eq:sigmawK-hat}]
\label{lem:unbiased}
Under Assumption~\ref{ass:sample-pi},
\[
  \E\big[\sigwhatsqplug\big]
  =\sigma_{w^{\star}}^2+\frac{K}{N}\Big(1-\textstyle\sum_j\pi_j^2\Big),
  \qquad
  \E\big[\sigwhatsq\big]=\sigma_{w^{\star}}^2 .
\]
The plug-in bias is non-negative, vanishing only in the degenerate case
$\sum_j\pi_j^2=1$ (all sampling mass in one cell), and grows with $K/N$;
the diagonal-removed estimator~\eqref{eq:sigmawK-hat} is exactly unbiased
for every $N\ge 2$ and every $K$.
\end{lemma}
\deferproof{unbiased}{%
\begin{proof}[Proof of Lemma~\ref{lem:unbiased}]
\begin{align}
  \mathrm{Bias}\big(\sigwhatsqplug\big)
  &=\E\Big[K\sum_{j=1}^{K}\hat\pi_j^2-1\Big]
    -\Big(K\sum_{j=1}^{K}\pi_j^2-1\Big)
   =K\sum_{j=1}^{K}\Big(\E\big[\hat\pi_j^2\big]-\pi_j^2\Big), \\
  \mathrm{Bias}\big(\sigwhatsq\big)
  &=\E\Big[K\sum_{j=1}^{K}\frac{n_j(n_j-1)}{N(N-1)}-1\Big]
    -\Big(K\sum_{j=1}^{K}\pi_j^2-1\Big)
   =K\sum_{j=1}^{K}\Big(\E\Big[\frac{n_j(n_j-1)}{N(N-1)}\Big]-\pi_j^2\Big).
  \label{eq:bias-diag}
\end{align}
It therefore suffices to evaluate the two cell-level expectations.

\emph{Step 1: moments of the counts.} Under
Assumption~\ref{ass:sample-pi} the counts are multinomial, so each
marginal is $n_j\sim\mathrm{Binomial}(N,\pi_j)$, giving
\[
  \E[n_j]=N\pi_j,\qquad \Var(n_j)=N\pi_j(1-\pi_j),\qquad
  \E[n_j^2]=\Var(n_j)+\E[n_j]^2=N\pi_j(1-\pi_j)+N^2\pi_j^2 .
\]

\emph{Step 2: the plug-in cell term is biased.} Dividing the second
moment by $N^2$,
\[
  \E\big[\hat\pi_j^2\big]=\frac{\E[n_j^2]}{N^2}
  =\pi_j^2+\frac{1}{N}\,\pi_j(1-\pi_j)
  \;>\;\pi_j^2 \quad\text{whenever }0<\pi_j<1 ,
\]
so each cell term over-estimates $\pi_j^2$ by $\pi_j(1-\pi_j)/N$.
Substituting into \eqref{eq:bias-plug} and using $\sum_j\pi_j=1$,
\[
  \mathrm{Bias}\big(\sigwhatsqplug\big)
  =K\sum_{j=1}^{K}\frac{\pi_j(1-\pi_j)}{N}
  =\frac{K}{N}\Big(\sum_{j=1}^{K}\pi_j-\sum_{j=1}^{K}\pi_j^2\Big)
  =\frac{K}{N}\Big(1-\sum_{j=1}^{K}\pi_j^2\Big),
\]
which is the first claim. It is non-negative because
$\sum_j\pi_j^2\le\big(\sum_j\pi_j\big)^2=1$, vanishing only when all
sampling mass lies in a single cell, and it grows linearly in $K/N$.

\emph{Step 3: removing the self-pairs.} The inflation arises because
$n_j^2$ counts all $N^2$ \emph{ordered} pairs of profiles falling in
$C_j$, including the $N$ pairs of a profile with itself. A pair of
\emph{distinct} profiles lies in $C_j$ with probability $\pi_j^2$, since
the two are independent draws (Assumption~\ref{ass:sample-pi}); a
profile paired with itself lies there with probability $\pi_j$, as it
coincides with itself by definition. The self-pairs are thus spurious
coincidences, and $n_j(n_j-1)=n_j^2-n_j$ removes exactly those $N$
terms, leaving the $N(N-1)$ ordered pairs of distinct profiles. Hence
\[
  \E\big[n_j(n_j-1)\big]=N(N-1)\,\pi_j^2 ,
  \qquad\text{so}\qquad
  \E\Big[\frac{n_j(n_j-1)}{N(N-1)}\Big]=\pi_j^2 ,
\]
the normalization matching the number of pairs retained. Each cell term
is therefore exactly unbiased for $\pi_j^2$, and \eqref{eq:bias-diag}
gives
\[
  \mathrm{Bias}\big(\sigwhatsq\big)
  =K\sum_{j=1}^{K}\big(\pi_j^2-\pi_j^2\big)=0 ,
\]
i.e. $\E\big[\sigwhatsq\big]=\sigma_{w^{\star}}^2$
for every $N\ge2$ and every $K$.
\end{proof}%
}


\begin{remark}[Trajectory dependence and the between-float estimator]
\label{rem:trajectory}
Estimator~\eqref{eq:sigmawK-hat} treats all profiles as independent draws
from $\pi$ (Assumption~\ref{ass:sample-pi}), yet profiles from one float
trace a single advective path and are not mutually independent. The
resulting error is nonetheless one-sided under the physically relevant
form of that dependence. Writing $c_a$ for the cell of profile $a$ and
extending the computation of Lemma~\ref{lem:unbiased} without its
independence step,
\begin{equation}
  \E\big[\sigwhatsq\big]-\sigma_{w^{\star}}^2
  =\frac{K}{N(N-1)}\sum_{a\ne b}
     \Big(\Prob(c_a=c_b)-\textstyle\sum_j\pi_j^2\Big),
  \label{eq:diag-bias-identity}
\end{equation}
an exact identity over the ordered pairs of distinct profiles. Two
independent draws from $\pi$ share a cell with probability exactly
$\sum_j\pi_j^2$, so each summand is the excess coincidence of a pair over
that baseline. Between-float pairs contribute nothing when distinct floats
are independent, and a non-negative amount under the shared advection
discussed below; within-float pairs---two cycles of one float, some ten
days and a short displacement apart relative to the cell scale---fall in a
common cell at least as often as independent draws, so their summands are
non-negative. Hence $\E\big[\sigwhatsq\big]\ge\sigma_{w^{\star}}^2$:
within-float clustering inflates \eqref{eq:sigmawK-hat}, the conservative
direction for an upper bound on $\sigma_{w^{\star}}$.
\end{remark}

Therefore, we also present an alternative \emph{between-float} estimator that counts only coincidence pairs from distinct floats.
Let $m_{\ell j}$ be the number of profiles of float $\ell$ in cell $j$ and $d_\ell=\sum_j m_{\ell j}$ its total, \begin{equation} \sigwhatsqbf = K\,\frac{\sum_j n_j(n_j-1)-\sum_\ell\sum_j m_{\ell j}(m_{\ell j}-1)} {N(N-1)-\sum_\ell d_\ell(d_\ell-1)}\;-\;1 , \label{eq:sigmawK-bf} \end{equation} which removes the within-float diagonal and normalizes by the between-float pair total.
Under the assumption that \emph{distinct} floats are independent draws from $\pi$, it is unbiased for $\sigma_{w^{\star}}^2$ (Lemma~\ref{lem:bf-unbiased}), with no within-float independence required, so it removes the within-float autocorrelation to which \eqref{eq:sigmawK-hat} is exposed.
This between-float independence is, however, only approximate: floats with a shared deployment or advective history---nearby releases entrained in the same current or mesoscale features---have positively associated cell memberships, so two of their profiles fall in a common cell with probability at least $\sum_j\pi_j^2$, and hence $\E\big[\sigwhatsqbf\big]\ge\sigma_{w^{\star}}^2$. The residual bias is thus in the same upward direction as the within-float clustering the estimator removes---conservative for an upper bound on $\sigma_{w^{\star}}$---and the between-float coherence it leaves is the weaker term, the same water parcel no longer being shared and trajectories diverging over time. We therefore use \eqref{eq:sigmawK-bf} in this spirit, as a conservative input to the bound rather than an exactly unbiased estimator.
Comparing \eqref{eq:sigmawK-hat} and \eqref{eq:sigmawK-bf}, which differ only through the within-float diagonal terms removed from the latter, is moreover an assumption-free diagnostic of how much of the sampling variability is trajectory autocorrelation rather than preferential sampling.

\begin{lemma}[Unbiasedness of the between-float estimator]
\label{lem:bf-unbiased}
If profiles from \emph{distinct} floats are independent draws with marginal
distribution $\pi$ then $\E\big[\sigwhatsqbf\big] = \sigma_{w^{\star}}^2$.
\end{lemma}
\deferproof{bf-unbiased}{%
\begin{proof}[Proof of Lemma~\ref{lem:bf-unbiased}]
Index profiles by $(\ell,i)$, with $\ell=1,\dots,L$ the float and
$i=1,\dots,d_\ell$ the profile within float $\ell$; let
$c_{\ell i}\in\{1,\dots,K\}$ be the cell membership of profile $(\ell,i)$, so that
\[
  m_{\ell j}=\sum_{i=1}^{d_\ell}\1\{c_{\ell i}=j\},\qquad
  n_j=\sum_{\ell}m_{\ell j},\qquad
  N=\sum_\ell d_\ell .
\]
We use only the two hypotheses of the lemma:
\begin{enumerate}
\item[(A1)] $\Prob(c_{\ell i}=j)=\pi_j$ for every profile $(\ell,i)$ and
  cell $j$;
\item[(A2)] the cell membership vectors of distinct floats are independent,
  $(c_{\ell i})_{i\le d_\ell}\perp\!\!\!\perp(c_{\ell'i'})_{i'\le d_{\ell'}}$
  for $\ell\ne\ell'$.
\end{enumerate}
Nothing is assumed about the joint distribution
$(c_{\ell 1},\dots,c_{\ell d_\ell})$ of the cells within a single float.
Write
$S_{\mathrm{bf}}=\sum_j n_j(n_j-1)-\sum_\ell\sum_j m_{\ell j}(m_{\ell j}-1)$
and $P_{\mathrm{bf}}=N(N-1)-\sum_\ell d_\ell(d_\ell-1)$ for the numerator
and denominator of~\eqref{eq:sigmawK-bf}, so
$\sigwhatsqbf=K\,S_{\mathrm{bf}}/P_{\mathrm{bf}}-1$.

The whole argument rests on one structural observation: $S_{\mathrm{bf}}$
and $P_{\mathrm{bf}}$ are sums over the \emph{same} index set, the ordered
pairs of profiles drawn from two \emph{distinct} floats. The denominator
counts all such pairs; the numerator counts those that land in a common
cell. Each pair coincides with one and the same expected probability
$\sum_j\pi_j^2$, so $\E[S_{\mathrm{bf}}]=\bigl(\sum_j\pi_j^2\bigr)
P_{\mathrm{bf}}$ term by term, and the ratio is exactly $\sum_j\pi_j^2$.
The three steps below make this precise.

\emph{Step 1 (the between-float pair set).}
Let
\[
  \mathcal B=\bigl\{\bigl((\ell,i),(\ell',i')\bigr):\ \ell\ne\ell',\
    i\le d_\ell,\ i'\le d_{\ell'}\bigr\}
\]
be the ordered pairs of profiles belonging to two distinct floats. Its
size depends only on the design $(d_\ell)_\ell$: since $N=\sum_\ell d_\ell$,
\[
  |\mathcal B|=\sum_{\ell\ne\ell'}d_\ell d_{\ell'}
  =\Bigl(\sum_\ell d_\ell\Bigr)^2-\sum_\ell d_\ell^2
  =N(N-1)-\sum_\ell d_\ell(d_\ell-1)=P_{\mathrm{bf}} .
\]
Thus $P_{\mathrm{bf}}$ is a deterministic constant, strictly positive as
soon as two floats each contribute a profile.

\emph{Step 2 ($S_{\mathrm{bf}}$ counts the coinciding pairs in $\mathcal B$).}
Expanding $n_j=\sum_\ell m_{\ell j}$ separates diagonal from off-diagonal
terms,
\[
  n_j(n_j-1)=n_j^2-n_j
  =\sum_\ell m_{\ell j}(m_{\ell j}-1)+\sum_{\ell\ne\ell'}m_{\ell j}m_{\ell'j},
\]
so summing over $j$ cancels the within-float diagonal against the term
subtracted in $S_{\mathrm{bf}}$, leaving
\[
  S_{\mathrm{bf}}=\sum_{\ell\ne\ell'}\sum_{j=1}^{K}m_{\ell j}m_{\ell'j}
  =\sum_{\ell\ne\ell'}\sum_{i=1}^{d_\ell}\sum_{i'=1}^{d_{\ell'}}
    \sum_{j=1}^{K}\1\{c_{\ell i}=j\}\1\{c_{\ell'i'}=j\}
  =\sum_{((\ell,i),(\ell',i'))\in\mathcal B}\1\{c_{\ell i}=c_{\ell'i'}\},
\]
where the second equality inserts the definition of $m_{\ell j}$ and the
third uses
$\sum_j\1\{c_{\ell i}=j\}\1\{c_{\ell'i'}=j\}=\1\{c_{\ell i}=c_{\ell'i'}\}$. So
$S_{\mathrm{bf}}$ is exactly the number of pairs in $\mathcal B$ that fall
in a common cell; every pair with $\ell=\ell'$ has dropped out.

\emph{Step 3 (each pair in $\mathcal B$ coincides with probability
$\sum_j\pi_j^2$).}
Fix $\bigl((\ell,i),(\ell',i')\bigr)\in\mathcal B$, so $\ell\ne\ell'$.
Decomposing over the common cell and applying (A2) then (A1),
\[
  \E\bigl[\1\{c_{\ell i}=c_{\ell'i'}\}\bigr]
  =\sum_{j=1}^{K}\Prob\bigl(c_{\ell i}=j,\;c_{\ell'i'}=j\bigr)
  \overset{\text{(A2)}}{=}\sum_{j=1}^{K}\Prob(c_{\ell i}=j)\Prob(c_{\ell'i'}=j)
  \overset{\text{(A1)}}{=}\sum_{j=1}^{K}\pi_j^{2}.
\]
This is the same constant for every pair in $\mathcal B$, independent of
$i,i'$; (A2) is invoked only across the two distinct floats $\ell\ne\ell'$,
never within a float.

\emph{Step 4 (assembly).}
By Step 2 and linearity, summing the constant of Step 3 over the
$|\mathcal B|=P_{\mathrm{bf}}$ pairs of Step 1,
\[
  \E[S_{\mathrm{bf}}]
  =\sum_{((\ell,i),(\ell',i'))\in\mathcal B}
     \E\bigl[\1\{c_{\ell i}=c_{\ell'i'}\}\bigr]
  =P_{\mathrm{bf}}\sum_{j=1}^{K}\pi_j^{2}.
\]
Since $P_{\mathrm{bf}}$ is deterministic and nonzero,
\[
  \E\bigl[\sigwhatsqbf\bigr]
  =\frac{K\,\E[S_{\mathrm{bf}}]}{P_{\mathrm{bf}}}-1
  =K\sum_{j=1}^{K}\pi_j^{2}-1
  =\sigma_{w^{\star}}^{2}.
\]
The within-float contributions $\sum_\ell\sum_j m_{\ell j}(m_{\ell j}-1)$ and
$\sum_\ell d_\ell(d_\ell-1)$ cancelled in Steps~1--2, before any expectation was
taken, so no assumption on the dependence among a single float's profiles
was needed. (Such dependence still affects $\Var[\sigwhatsqbf]$, but not
its mean.)
\end{proof}%
}

\subsection{Testing uniform sampling}
\label{sec:uniformity-test}

The dispersion $\sigwhat$ also supports a formal test of the null of
uniform sampling, $H_0:\pi=\nu$ (equivalently $\pi_j=1/K$ for all $j$).
Under $H_0$ the plug-in dispersion is a rescaled Pearson
goodness-of-fit statistic,
\[
  X^2=\sum_{j=1}^{K}\frac{(n_j-N/K)^2}{N/K}
     =N\,\sigwhatsqplug,
\]
and preferential sampling makes $X^2$ (and $\sigwhatsq$)
large. The classical reference $X^2\sim\chi^2_{K-1}$ is \emph{not} valid
here: the profiles are cluster-dependent (Remark~\ref{rem:trajectory}),
which over-disperses the cell counts and inflates $X^2$ under $H_0$, so a
$\chi^2_{K-1}$ $p$-value over-rejects and reports spurious preferential
sampling; the sparse expected counts $N/K$ compound this.

We calibrate instead by a float-level permutation null that preserves
the within-float trajectory dependence. Each whole float is given an
independent uniform rigid shift in the reference coordinates
$(\mathrm{lon},\ \sin\mathrm{lat},\ t)$, wrapped toroidally onto the
domain; this leaves each float's internal geometry---hence its
within-float clustering and $d_\ell$---intact while placing floats
independently and uniformly, so the permutation distribution reproduces
the within-float over-dispersion of the data under $H_0$. It does not
reproduce \emph{between}-float dependence: the independent per-float
shifts place floats independently, whereas shared advection positively
correlates nearby floats, so the null is mildly under-dispersed and the
test correspondingly anti-conservative in that component. Removing this
would require shifting whole advective clusters jointly, which we leave
to future work; here we simply flag the direction of the effect. For a statistic
$T\in\{X^2,\ \sigwhatsqbf\}$ with observed
value $T_{\mathrm{obs}}$ and $B$ permutation replicates
$T^{(1)},\dots,T^{(B)}$, the one-sided $p$-value is
\[
  \hat p=\frac{1+\#\{b:\,T^{(b)}\ge T_{\mathrm{obs}}\}}{B+1}.
\]
The between-float statistic~\eqref{eq:sigmawK-bf} is the more
informative choice: it is free of the within-float clustering that the
permutation must otherwise absorb, and is centered at $0$ under $H_0$
when distinct floats are independent (Lemma~\ref{lem:bf-unbiased}). It
does not remove between-float dependence, which the per-float shift also
omits from the null, so a small $\hat p$ is evidence against uniformity
\emph{or} against unmodelled between-float coherence, and should be read
with the caveat above.

Two cautions. A large $\hat p$ is evidence of \emph{no detectable}
departure from uniform, not of exact uniformity; with few floats the
test has limited power, so we treat it as corroboration and let the
effect size $\sigwhat$ with its interval---in particular the upper
confidence limit the bound requires---carry the quantitative claim.
Second, the gap between the permutation $\hat p$ and the naive
$\chi^2_{K-1}$ $p$-value is itself worth reporting, as it quantifies how
much the trajectory dependence would have inflated a naive test.

\subsection{Estimating $\sigma_{Y^{\star}}$ and $\sigma_{Y^{\perp}}$ from a
reanalysis}
\label{sec:sigmaY}

The field factors require a value of $y$ throughout $\X$, including the
space--time cells the floats avoided. We take it from an independent
reanalysis (GLORYS12; Remark~\ref{rem:glorys}) rather than from the
float observations, and the reason is not convenience but bias.

\begin{proposition}[Float-only field dispersion is contaminated]
\label{prop:float-contam}
The empirical dispersion of $Y$ over the float locations estimates
$\Var_\pi(Y)$, not $\Var_\nu(Y)=\sigma_Y^2$, and
\[
  \Var_\pi(Y)-\Var_\nu(Y)=\Cov_\nu\!\big(w,(Y-\mu)^2\big).
\]
\end{proposition}
\deferproof{float-contam}{%
\begin{proof}[Proof of Proposition~\ref{prop:float-contam}]
The float locations are draws from $\pi$ (Assumption~\ref{ass:sample-pi}),
so their empirical variance is consistent for $\Var_\pi(Y)$. Using
$\dd\pi=w\dd\nu$ and $\E_\nu[w]=1$,
$\Var_\pi(Y)=\int(y-\mu_\pi)^2 w\dd\nu$; expanding and comparing with
$\int(y-\mu)^2\dd\nu$ gives the stated covariance, the sampling bias of
the squared-anomaly field.
\end{proof}%
}

The discrepancy is non-zero for exactly the reason $B_{\mathrm{samp}}$
is: it is the $w$--$Y$ coupling the paper studies, applied to
$(Y-\mu)^2$. A float-only $\hat\sigma_Y$ would thus be inflated when
floats favor the high-variability front and deflated when they collect
in quiescent water. De-biasing by importance weighting
($\widehat{\sigma_Y^2}=\frac1n\sum_i w(x_i)^{-1}(y(x_i)-\hat\mu)^2$)
recovers $\Var_\nu(Y)$ in principle but requires $w$---the object we
avoid modeling---and is dominated by $1/w$ in the undersampled cells
where floats are scarcest, so its variance is largest exactly where the
bound is most sensitive. We therefore do not estimate $\sigma_Y$ from
the floats.

The reanalysis provides $y$ over $\nu$: constrained by satellite
altimetry, sea-surface temperature, and model dynamics, it carries an
OHC value in every cell, including the float-sparse ones, breaking the
dependence on the Argo sampling pattern. On the level-$K$ partition,
with $y$ evaluated on the native $1/12^\circ$ grid and aggregated to
cells,
\begin{align}
  \sigYhatsq
  &=\sum_{j=1}^{K}\tfrac1K\big(\bar y_j-\hat\mu\big)^2,
  &\bar y_j&=\text{area-weighted mean of $y$ over the grid nodes in }C_j,
  \ \hat\mu=\tfrac1K\textstyle\sum_j\bar y_j,
  \label{eq:sigmaYK-hat}\\[2pt]
  \sigYperphatsq
  &=\sum_{j=1}^{K}\tfrac1K\,\widehat{\Var}_\nu\!\big(y\mid C_j\big),
  &&\text{(mean within-cell variance of $y$).}
  \label{eq:sigmaYperp-hat}
\end{align}
These are direct quadratures of a densely observed field, so---unlike
$\sigwhatsq$---no diagonal removal or count-noise correction is
needed, and $\sigma_Y^2=\sigma_{Y^{\star}}^2+\sigma_{Y^{\perp}}^2$ is
recoverable in full. Computing $\sigYperphat$ from the native grid
rather than pre-averaging to cells is essential: it is the genuine
sub-cell field variance, and it is the quantity that controls the
unestimated residual in Proposition~\ref{prop:scale-decomp}.

\begin{remark}[Direction of reanalysis error]
\label{rem:conservative-Y}
Data assimilation tends to \emph{attenuate} anomaly amplitude, so a
reanalysis is liable to understate $\sigma_Y$---the unsafe direction for
a bound. We treat the field proxy as a sensitivity axis rather than a
single source: $\sigYhat$ and $\sigYperphat$ are recomputed on
at least one product with different Argo dependence (e.g. an
Argo-independent or differently-assimilated analysis), and the
\emph{largest} $\hat\sigma_Y$ across products is carried into the bound.
Agreement across products with differing Argo reliance also answers,
empirically, the concern that the reanalysis merely re-expresses the
Argo data (Remark~\ref{rem:glorys}).
\end{remark}


\subsection{Uncertainty quantification}
\label{sec:uncertainty}

The two factors carry different kinds of uncertainty, and we keep them
separate.

For $\sigwhat$, the relevant variability is sampling variability
over floats, and the multinomial/normal approximation to the $\chi^2$
statistic is invalid because of the trajectory dependence of
Remark~\ref{rem:trajectory}. Because
$\sum_j n_j(n_j-1)=\sum_{\ell\neq\ell'}O_{\ell\ell'}$, where
$O_{\ell\ell'}$ is the footprint overlap of floats $\ell$ and $\ell'$,
the estimator is a degree-two U-statistic in the floats; a
with-replacement float bootstrap is consequently biased, since a
duplicated float injects a spurious self-overlap $O_{\ell\ell}$. We
therefore use a \emph{delete-one-float jackknife}: recompute the
estimator with each float's profiles removed in turn and form the
jackknife variance over the $L=40$ floats. The
float---not the profile---is the resampling unit, which reflects the
effective sample size and propagates the within-float correlation into
the interval; the intervals widen at the fine end of the refinement
curve, honestly displaying what $40$ tracks can resolve.

Concretely, both estimators are functions of the $L$ float footprints
$\{m_{\ell j}\}$, and the numerator of \eqref{eq:sigmawK-hat} decomposes
over float pairs as
$\sum_j n_j(n_j-1)=\sum_{\ell\ne\ell'}O_{\ell\ell'}+\sum_\ell\sum_j m_{\ell j}(m_{\ell j}-1)$
with $O_{\ell\ell'}=\sum_j m_{\ell j}m_{\ell'j}$, exhibiting it as a
degree-two $U$-statistic in the floats. The delete-one-float jackknife
variance is
\begin{equation}
  \widehat V_{\mathrm{jack}}
  =\frac{L-1}{L}\sum_{\ell=1}^{L}\big(\hat\theta_{(-\ell)}-\bar\theta_{(\cdot)}\big)^2,
  \qquad \bar\theta_{(\cdot)}=\frac1L\sum_{\ell=1}^{L}\hat\theta_{(-\ell)},
  \label{eq:jackvar}
\end{equation}
where $\hat\theta_{(-\ell)}$ recomputes $\sigwhatsq$ (or its
between-float variant) with float $\ell$'s profiles removed; the reported
interval is $\hat\theta\pm z_{1-\alpha/2}\,\widehat V_{\mathrm{jack}}^{1/2}$
on the $\sigma^2$ scale, with $\sigma$-scale endpoints
$\sqrt{\max(0,\cdot)}$. For $U$-statistics this jackknife is consistent
for the asymptotic variance \citep{arvesen1969}.

The with-replacement float bootstrap, by contrast, is biased for this
functional. When a float is resampled more than once, its identical
copies form coincidence pairs that two distinct floats never would, each
adding a spurious self-overlap $O_{\ell\ell}=\sum_j m_{\ell j}^2$; these ties shift
the bootstrap distribution of $\sigwhatsq$ upward and break
interval coverage---in simulation a with-replacement float bootstrap
attained $\approx 50\%$ coverage against a nominal $95\%$, whereas the
jackknife \eqref{eq:jackvar} attained $\approx 95\%$. The jackknife only
deletes and never duplicates, so it avoids the ties; we use it
throughout and retain the bootstrap only as a diagnostic.

For $\sigma_Y$, the relevant uncertainty is field-proxy uncertainty,
not sampling uncertainty, so a float bootstrap is inappropriate. We
quantify it by the spread of $\hat\sigma_Y$ across reanalysis products
(Remark~\ref{rem:conservative-Y}), carrying the conservative (largest)
value and reporting the range.

The reported bound at each resolution is
$\sigwhat\,\sigYhat$, with an interval obtained by
combining the delete-one-float jackknife interval \eqref{eq:jackvar} of
$\sigwhat$ with the across-product range of $\sigYhat$,
accompanied by the
residual diagnostic $\sigYperphat/\hat\sigma_Y$. To interpret
severity, we express the bound relative to an external yardstick---the
estimated regional OHC trend over the period, and the posterior
standard deviation of the hierarchical OHC model
(Remark~\ref{rem:coherence})---so that ``not severe'' is stated as a
fraction of a signal the reader already trusts rather than in absolute
units.

\subsection{Summary of the procedure}
\label{sec:procedure}

\begin{enumerate}
\item Fix the relevance scale from an empirical OHC variogram on the
  reanalysis; choose a bracket of equal-measure partitions
  (Definition~\ref{def:partition}) spanning the ${\approx}25$--$76$\,km
  sampling band \eqref{eq:cell-km}, verifying it is at or above the
  relevance scale.
\item Retain all profiles (no thinning): the within-float clustering
  this leaves biases $\sigwhat$ upward only, the safe direction
  for the bound (Remark~\ref{rem:trajectory}).
\item At each partition: form counts $\{n_j\}$ and compute
  $\sigwhat$ by \eqref{eq:sigmawK-hat}; aggregate the
  $1/12^\circ$ reanalysis to the same cells and compute
  $\sigYhat$ and $\sigYperphat$ by
  \eqref{eq:sigmaYK-hat}--\eqref{eq:sigmaYperp-hat}.
\item Repeat step~3 across reanalysis products; retain the largest
  $\sigYhat$ (Remark~\ref{rem:conservative-Y}).
\item Delete-one-float jackknife $\sigwhat$ for intervals
  \eqref{eq:jackvar} (Section~\ref{sec:uncertainty}).
\item Report the refinement curve of the bound
  $\sigwhat\sigYhat$ with intervals, the residual
  diagnostic $\sigYperphat/\hat\sigma_Y$, and the bound as a
  fraction of the OHC trend and of the model posterior SD.
\end{enumerate}

The resulting guarantee depends on the data through counts of floats in
cells and averages of an independent field over those cells; the only
modeling choice is the cell scale, which is pinned to a measured
decorrelation length and displayed as a curve rather than fixed
silently. The complex spatio-temporal intensity model of
Remark~\ref{rem:hat-w} and the advection simulator are reserved as
independent cross-checks on this estimate, not as inputs to it.
