---
title: "argo bias"
subtitle: "exploratory analysis notebook"
author : daniel ross
date-modified: last-modified
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    theme: default
    toc: true
    toc-depth: 6
    toc-title: "contents"
    toc-location: right
    number-sections: true
    code-fold: true
    code-tools: true
lightbox: true
execute:
  cache: true
  freeze: auto
  warning: false
  message: false
  prompt: true
  fig-align: center
  fig-width: 11
  fig-height: 8.5
params:
  archive_eval: false
knitr:
  opts_chunk:
    prompt: false
    collapse: false
editor_options:
  chunk_output_type: console
---

# Scratch {.unnumbered}

::: {.callout-note collapse="true"}
# 2026-06-10

```{r}
#| eval: !expr params$archive_eval
track_cohort <- function(df, yrs = NULL, n, f = map_bin, scale.limits = NULL, plot.ncol = 2, ...){
  if(is.null(yrs)) yrs <- df[,unique(year)]
  fid_new_yr <- df[(year %in% yrs) & (float_obs_n == 1), unique(float_id)]
  A <- df[float_id %in% fid_new_yr]
  plots <- list()
  for(i in seq_along(n)){
    p <- f(A[float_obs_n == n[i]], ...) +
      scale_fill_viridis_c(limits = scale.limits) +
      labs(title = paste("obs", n[i]))
    plots[[i]] <- p
  }
  wrap_plots(plots, ncol = plot.ncol, byrow = FALSE, guides = "collect")
  }
```

# float concentration

## (new) floats by year
```{r}
#| eval: !expr params$archive_eval
new_floats_yr <- argo_ohc[,.(
  floats = uniqueN(float_id),
  floats_new = sum(float_obs_n == 1)
), keyby = c("year")
]
new_floats_month <- argo_ohc[,.(
  floats = uniqueN(float_id),
  floats_new = sum(float_obs_n == 1)
), keyby = c("year", "month")
]
obs_p_float <- argo_ohc[,.(
  obs = .N
),float_id]
summary(obs_p_float$obs)
hist(obs_p_float$obs, breaks = 100)
```

## track cohorts of new floats
```{r}
#| eval: !expr params$archive_eval
new_floats_month[year == 2005] 
fid_new_2005 <- argo_ohc[(year == 2005) & (float_obs_n == 1), unique(float_id)] 
length(fid_new_2005)

A <- argo_ohc[float_id %in% fid_new_2005]
A[, .(
  floats = uniqueN(float_id),
  obs = .N
), keyby = year]

obs_p_float <- A[,.(
  obs = .N
),float_id]
summary(obs_p_float$obs)
hist(obs_p_float$obs, breaks = seq(0,510,10))
```

```{r}
#| eval: !expr params$archive_eval
V <- argo_ohc[!is.na(speed_ms) & speed_ms <= 0.5,]
```

```{r}
#| eval: !expr params$archive_eval
VO <- argo_ohc[!is.na(speed_ms) & speed_ms > 0.5,] 
```

```{r}
#| eval: !expr params$archive_eval
A <- argo_nac[!(float_all_dt_day_10) & !(dt_day_10)]
summary(A$dt_day)
hist(A$dt_day)
hist(A[dt_day<100,dt_day])

summary(argo_nac$dt_day)

A <- argo_nac[(float_has_dt_day_10) & !is.na(dt_day), .N, dt_day]
summary(A$dt_day)
plot(x=A$dt_day, y=A$N)
hist(argo_nac[!is.na(dt_day) & (float_has_dt_day_10), dt_day], breaks = 100, main = "obs freq", xlab = NULL)
hist(argo_nac[!is.na(dt_day) & dt_day<31, dt_day], breaks = seq(0,31,.5), main = "obs freq", xlab = NULL)
```

```{r}
#| eval: !expr params$archive_eval
summary(argo_nac_clean$speed_ms)
```

```{r}
#| eval: !expr params$archive_eval
ggplot(argo_nac_clean, aes(x = u_ms, y = v_ms)) +
  stat_bin_2d(aes(fill = after_stat(density)), binwidth = 0.01) +
  scale_fill_viridis_c(option = "C") +
  labs(
    x = "u (m/s)",
    y = "v (m/s)",
    fill = "density"
  ) +
  theme_minimal() +
  coord_fixed()
```

```{r}
#| eval: !expr params$archive_eval
map_sum_pts(argo_nac_clean, z = "u_ms", xlim = any_nac_lon, ylim = any_nac_lat, binwidth = c(.25, .25))
```

```{r}
#| eval: !expr params$archive_eval
map_sum_pts(argo_nac_clean, z = "v_ms", xlim = any_nac_lon, ylim = any_nac_lat, binwidth = c(.25, .25))
```

```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,2))
hist(argo_nac_clean$ohc_700, breaks = 100)
hist(argo_nac_clean$ohc_2000, breaks = 100)
hist(argo_nac_clean$dohc_700, breaks = 100)
hist(argo_nac_clean$dohc_2000, breaks = 100)

summary(argo_nac_clean$dohc_700)
O <- argo_nac_clean[abs(dohc_700) > 1]
hist(log(argo_nac_clean$abs_dohc_700))
```

### velocity maps

- *averaged velocity within 1 deg by 1 deg cells across all years*

::: {.panel-tabset}

## eastward velocity
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(V, z = "u_ms") + labs(title = "u (m/s) eastward velocity")
```

## northward velocity
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(argo_ohc, z = "v_ms") +
  labs(title = "v (m/s) northward velocity")
```

## speed
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(argo_ohc, z = "speed_ms") + 
  labs(title = "speed (m/s)")
```

:::

:::

# Archive {.unnumbered}

::: {.callout-note collapse="true"}
# 2026-06-03

# literature
## float velocity
- [K. Lebedev, H. Yoshinari, N. A. Maximenko, and P. W. Hacker. YoMaHa'07: Velocity data assessed from trajectories of Argo floats at parking level and at the sea surface, IPRC Technical Note No. 4(2), June 12, 2007, 16p.](https://apdrc.soest.hawaii.edu/projects/yomaha/)
  - derive Argo drift speed at park depth to quantify ocean circulation at 1,000 m depth
  - See also, [Katsumata, K., Yoshinari, H. Uncertainties in global mapping of Argo drift data at the parking level. J Oceanogr 66, 553–569 (2010). https://doi.org/10.1007/s10872-010-0046-4](https://link.springer.com/article/10.1007/s10872-010-0046-4)
  - Estimate parking level (1000m) zonal (eastward $u$) and meridional (northward $v$) velocity components using the coordinates ($[X,Y]$ = [lon, lat]) of floats before it dives and after it re-arrives at the sea surface
    \begin{gather}
      (U_\text{deep}^n, V_\text{deep}^n) = ([X,Y]^n_\text{first} - [X,Y]^{n-1}_\text{last}) / (T^n_\text{first} - T^{n-1}_\text{last})
    \end{gather}
  - Use a regression to estimate surface velocity and errors for deep/surface velocities
- Above referenced in more recent literature, e.g., [Oke, Peter R., Tatiana Rykova, Gabriela S. Pilo, and Jennifer L. Lovell. "Estimating Argo float trajectories under ice." Earth and Space Science 9, no. 7 (2022): e2022EA002312.](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022EA002312) 

# eda

- *analysis based on preprocessed `argo_object_01.RData` and is therefore limited to float observations in January of each year from 2007 through 2016*
- *augmented `preprocess_argo_data.R` to include velocity calculations*

**[[Note, just began incorporating by-year anaylses so those are incomplete]]**

```{r}
#| eval: !expr params$archive_eval
options(digits = 4, scipen = 999)

library(tidyverse)
library(data.table)
library(geosphere)

load("../data/argo_velo_data_january.RData")
```

## float preliminaries

- *drop observations for floats with only one observation*
- **velocity was calculated within year, therefore the first observation for a float in each year will have a missing velocity**
  - e.g., the velocity is not calculated between a given float's 2015-01-31 and 2016-01-01 profile
```{r}
#| eval: !expr params$archive_eval
argo[,.(
  floats = uniqueN(float_id),
  obs = .N,
  obs_velo = sum(!is.na(speed_ms)),
  dates = paste0(range(isodatetime), collapse = " - ")
), float_r]

A <- argo[(float_r)]
```

```{r}
#| eval: !expr params$archive_eval
oceans <- sort(unique(A$ocean))
oceans <- oceans[oceans!="south china and easter archipelagic seas"]
```

### obs per float
::: {.panel-tabset}
## overall
```{r}
#| eval: !expr params$archive_eval
float_n <- A[,.N,float_id]
summary(float_n$N)
hist(float_n$N, breaks = seq(0,44,1), main = "obs per float", xlab = "n")
```

## region
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  float_n <- A[ocean == o,.N,float_id]
  hist(float_n$N, breaks = seq(0,44,1), main = o, xlab = "n")
}
```
:::

### surfacing/obs frequency

- *most floats do share data every 10 days*

```{r}
#| eval: !expr params$archive_eval
A[, dt_day := dt_s / (60^2 * 24)]
summary(A$dt_day)
B <- A[!is.na(dt_day), .N, dt_day]

par(mfrow = c(1,1))
hist(A$dt_day, breaks = seq(0,31,.05), main = "obs freq", xlab = NULL)
```

### ocean region 

- *for this analysis I used the global ocean regions*
- **[[update analysis to use finer ocean basin geographies. source?]]**
- *floats lat/lon reverse geocoded using shapefile available at <https://www.marineregions.org/sources.php#goas> and downloaded using the package `mregions2`*
- *dropped the 3 observations in the south china sea for regional analyses*


```{r}
#| eval: !expr params$archive_eval
float_loc <- A[,.(float_id, float_i, lon_degrees, lat_degrees, ocean, speed_ms)]

float_loc[,.(
  floats = uniqueN(float_id),
  obs = .N,
  obs_velo = sum(!is.na(speed_ms))
),ocean][order(-obs)]

sample_locs <- float_loc[, .SD[sample(.N, min(1000, .N))], by = ocean]

map_pts <- function(df, x = "lon_degrees", y = "lat_degrees", color = "ocean", size = .5, legend.rows = 1){
  ggplot() +
  borders("world", fill = "gray80", colour = "gray80") +
  geom_point(data = df,
    aes(x = .data[[x]],
        y = .data[[y]],
        color = .data[[color]]),
        size = size) +
  coord_sf() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = legend.rows))
}
map_sum_pts <- function(df, x = "lon_degrees", y = "lat_degrees", z = "u_ms", size = .5, legend.rows = 1){
  ggplot(df, aes(x = .data[[x]],y = .data[[y]],z = .data[[z]])) +
  borders("world", fill = "gray80", colour = "gray80") +
  stat_summary_2d(fun = mean, binwidth = c(1,1)) + 
  scale_fill_gradient2(midpoint = 0,low = "red",mid = "grey100",high = "blue") + 
  coord_sf() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_colorbar())
}

map_pts(sample_locs, color = "ocean", legend.rows = 3)
```

## velocity
- *velocity components u/v are calculated by finding the change in longitude/latitude and dividing by change in time.*
- *[`geosphere::distVincentyEllipsoid`](https://cran.r-project.org/web/packages/geosphere/refman/geosphere.html#distVincentyEllipsoid>), an ellipsoidal (WGS84 by default) great-circle-distance, is used to calculate the distance between float obs.*
- *also calculated the magnitude of the velocity (speed) and the bearing of the velocity vectors in radians*

### overall distributions
- *histograms of float velocity over entire ocean omitting extreme outliers corresponding to speeds >0.5m/s*
- *outliers are dropped for the remainder of this analysis*
```{r}
#| eval: !expr params$archive_eval
#| fig-height: 6
B <- A[!is.na(speed_ms)]

X <- A[!is.na(speed_ms) & speed_ms <= 0.5,]
X <- X[ocean != "south china and easter archipelagic seas"]
par(mfrow = c(1,3))
hist(X[,speed_ms], breaks = seq(0,.5,.01), main = "float speed", xlab = "speed (m/s)")
hist(X[,u_ms], breaks = seq(-.5,.5,.01), main = "float eastward velocity", xlab = "u (m/s)")
hist(X[,v_ms], breaks = seq(-.5,.5,.01), main = "float northward velocity", xlab = "v (m/s)")
```

```{r}
#| eval: !expr params$archive_eval
  ggplot(X, aes(x = u_ms, y = v_ms)) +
    stat_bin_2d(aes(fill = after_stat(density)), binwidth = 0.01) +
    scale_fill_viridis_c(option = "C") +
    labs(
      x = "u (m/s)",
      y = "v (m/s)",
      fill = "density"
    ) +
    theme_minimal() +
    coord_fixed()
```

```{r}
#| eval: !expr params$archive_eval
my.sum <- function(x) list(mean = mean(x), sd = sd(x))
X[, unlist(lapply(.SD, my.sum)), .SDcols = c("u_ms","v_ms","speed_ms")]
```

### outliers
- *for preliminary analyses simply dropped outliers and did not further analyze patterns in these obs*
```{r}
#| eval: !expr params$archive_eval
b <- c(0,.5,1,10,20,50,100,500,B[,max(speed_ms)])
B[, speed_g := cut(speed_ms, breaks = b, dig.lab = 5)]
B[,.(obs = .N),keyby = speed_g][,prop := obs / sum(obs)][, c_prop := cumsum(prop)][]
```

### velocity maps

- *averaged velocity within 1 deg by 1 deg cells across all years 2007-2016*

::: {.panel-tabset}

## eastward velocity
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(X, z = "u_ms") +
  labs(title = "u (m/s) eastward velocity")
```

## northward velocity
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(X, z = "v_ms") +
  labs(title = "v (m/s) northward velocity")
```

## speed
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(X, z = "speed_ms") + 
  labs(title = "speed (m/s)")
```

:::

### regional distributions

#### joint dist of velocity components
```{r}
#| eval: !expr params$archive_eval
ggplot(X, aes(x = u_ms, y = v_ms)) +
  stat_bin_2d(aes(fill = after_stat(density)), binwidth = 0.01) +
  facet_wrap(~ ocean, axes = "all") +
  scale_fill_viridis_c(option = "C") +
  labs(
    x = "u (m/s)",
    y = "v (m/s)",
    fill = "density"
  ) +
  theme_minimal() +
  coord_fixed()
```

```{r}
#| eval: !expr params$archive_eval
my.sum <- function(x) list(mean = mean(x), sd = sd(x))
X[, as.list(unlist(lapply(.SD, my.sum))), .SDcols = c("u_ms","v_ms","speed_ms"), keyby=ocean]
```

## temperature

- *histograms of centered temperature by profile pressure (~depth in m)*
- *0m and 2000m correspond to the min/max observed pressure for each float*
- *for other depths, found the temperature associated with the pressure obs closest to the target depth*
  - *e.g., the most common pressure (8.4% of floats) for the '1000m' observations is in fact 988m, while 7.7% of floats have an obs at exactly 1000.0m*
- **for temperature only analyses, used data filtered to floats with multiple observations; however, included the first profile observation for those floats for each year where we have missing velocity **

::: {.panel-tabset}
### 0m
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_0_c],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 0m"
  )
}
```
### 100m
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_100_c],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 100m"
  )
}
```
### 500m
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_500_c],
   # freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 500m"
  )
}
```
### 1000m
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_1000_c],
   # freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 1000m"
  )
}
```
### 2000m
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, temp_2000_c],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "centered temp 2000m"
  )
}
```
### vhc
```{r}
#| eval: !expr params$archive_eval
par(mfrow = c(2,4))
for (o in oceans) {
  hist(
    A[ocean == o, vhc_obs],
    #freq = FALSE,
    breaks = 30,
    main = o,
    xlab = "vhc"
  )
}
```
:::

### temperature maps

- *averaged temperature within 1 deg by 1 deg cells across all years 2007-2016*

::: {.panel-tabset}
## 0m
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(A, z = "temp_0") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_0)]) + 
  labs(title = "temp 0m")
```
## 100m
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(A, z = "temp_100") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_100)]) + 
  labs(title = "temp 100m")
```
## 500m
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(A, z = "temp_500") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_500)]) + 
  labs(title = "temp 500m")
```
## 1000m
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(A, z = "temp_1000") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_1000)]) + 
  labs(title = "temp 1000m")
```
## 2000m
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(A, z = "temp_2000") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(temp_2000)]) + 
  labs(title = "temp 2000m")
```
## vhc 
```{r}
#| eval: !expr params$archive_eval
map_sum_pts(A, z = "vhc_obs") + 
  scale_fill_gradient2(low = "blue",mid = "white",high = "red", midpoint = X[, mean(vhc_obs)]) + 
  labs(title = "vhc")
```
:::

## correlation between temperature and velocity

- *correlation between temperatures and velocities for floats with multiple observations within a year*
- *dtemp_ variables are the change in temperature for consecutive (within year) float observations at the given depth*
  - negligible correlations between change in temperature and velocity
- **small to moderate negative correlation between eastward velocity and temperature/vhc across depths in southern hemisphere**
  - southern antarctic ocean which has positive correlation between eastward velocity and temperature/vhc
- **negligible negative correlation between eastward velocity and temperature/vhc across depths in northern hemisphere**
  - artic ocean has larger correlations but few obs
- **negligible or no correlation between northward velocity and temperature**

```{r}
#| eval: !expr params$archive_eval
#| echo: false
velo_temp_cor <- function(df, digits = 4){
  cols <- c("u_ms", "v_ms", "speed_ms",paste0("temp_",c(0,500,1000,2000)),"vhc_obs",paste0("dtemp_",c(0,500,1000,2000)),"dvhc_obs")
  V <- df[,..cols]
         return(round(cor(V)[-c(1:3),1:3],digits))
}
```

### excluding outlier velocities
#### overall
```{r}
#| eval: !expr params$archive_eval
#| results: asis
#| echo: false
cat("::: {.panel-tabset}\n\n")

cat("## overall", "\n\n")
tab <- knitr::kable(velo_temp_cor(X))
cat(tab, sep = "\n")
cat("N:", X[,.N], sep = "\n")
cat("\n")

for (o in oceans) {
  cat("##", o, "\n\n")
  tab <- knitr::kable(velo_temp_cor(X[ocean == o,]))
  cat(tab, sep = "\n")
  cat("N:", X[ocean == o,.N], sep = "\n")
  cat("\n")
  cat("\n\n")
}

cat(":::\n")
```

#### by year
::: {.panel-tabset}

```{r}
#| eval: !expr params$archive_eval
#| results: asis
#| echo: false

for (yr in 2007:2016){
  cat("##", yr, "\n\n")
  cat("::: {.panel-tabset}\n\n")

  cat("## overall", "\n\n")
  Y <- X[years == yr]
  tab <- knitr::kable(velo_temp_cor(Y))
  cat(tab, sep = "\n")
  cat("N:", Y[,.N], sep = "\n")
  cat("\n")

  for (o in oceans) {
    cat("##", o, "\n\n")
    tab <- knitr::kable(velo_temp_cor(Y[ocean == o,]))
    cat(tab, sep = "\n")
    cat("N:", Y[ocean == o,.N], sep = "\n")
    cat("\n")
    cat("\n\n")
  }
  cat(":::\n")
}
```

:::

### including outlier velocities

#### overall
- *overall, south atlantic, and southern correlations are sensitive to outliers*

```{r}
#| eval: !expr params$archive_eval
#| results: asis
#| echo: false

Y <- A[!is.na(speed_ms)]

cat("::: {.panel-tabset}\n\n")

cat("## overall", "\n\n")
tab <- knitr::kable(velo_temp_cor(Y))
cat(tab, sep = "\n")
cat("N:", Y[,.N], sep = "\n")
cat("\n")

for (o in oceans) {
  cat("##", o, "\n\n")
  tab <- knitr::kable(velo_temp_cor(Y[ocean == o,]))
  cat(tab, sep = "\n")
  cat("N:", Y[ocean == o,.N], sep = "\n")
  cat("\n")
  cat("\n\n")
}

cat(":::\n")
```

### heatmaps
- *centered temperatures (across all years) averaged within 0.01m/s cells of (u, v)*
- *positive eastward velocity appears to be correlated with negative temperatures*

```{r}
#| eval: !expr params$archive_eval
plot_velocity_heatmap <- function(df,z,title = z,binwidth = 0.01, midpoint = 0, ocean = T) {

  p <- ggplot(df, aes(x = u_ms, y = v_ms, z = .data[[z]])) +
    stat_summary_2d(fun = mean, binwidth = binwidth) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
      midpoint = midpoint, name = "temp") +
    labs(x = "u (m/s)",y = "v (m/s)",title = title) +
    theme_minimal() +
    theme(legend.position = "bottom") +
    coord_equal()

  if(ocean){
    p <- p + facet_wrap(~ocean)
  }
  p
}
```

#### overall
::: {.panel-tabset}
## 0m
```{r}
#| eval: !expr params$archive_eval
plot_velocity_heatmap(X, "temp_0_c", "centered ocean temp at 0m")
```
## 100m
```{r}
#| eval: !expr params$archive_eval
plot_velocity_heatmap(X, "temp_100_c", "centered ocean temp at 100m")
```
## 500m
```{r}
#| eval: !expr params$archive_eval
plot_velocity_heatmap(X, "temp_500_c", "centered ocean temp at 500m")
```
## 1000m
```{r}
#| eval: !expr params$archive_eval
plot_velocity_heatmap(X, "temp_1000_c", "centered ocean temp at 1000m")
```
## 2000m
```{r}
#| eval: !expr params$archive_eval
plot_velocity_heatmap(X, "temp_2000_c", "centered ocean temp at 2000m")
```
:::

:::
